---
title: "medicalcoder"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{medicalcoder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(medicalcoder)
```

## Quick start

The essentials for first-time users:

* Input must be **long**: one ICD code per row.
* Required columns for `comorbidities()`:
  - `id.vars`: one or more IDs (e.g., `patid`, `encounter`).
  - `icd.codes`: column with the ICD code (full or compact).
  - `icdv`: 9/10 for ICD-9/10.
  - `dx`: 1 = diagnosis, 0 = procedure.
* Present-on-admission (POA): either a single value (`poa = 0/1`) or a column (`poa.var`).
* Primary diagnosis: excluded by default (`primarydx = 0`); include via `primarydx = 1` or `primarydx.var`.

### Minimal Elixhauser (AHRQ 2026, ICD-10)

Use the bundled `mdcr` data in long format.

```{r elixhauser}
data(mdcr, package = "medicalcoder")

elix_sample <-
  comorbidities(
    data      = mdcr,
    id.vars   = "patid",
    icd.codes = "code",
    icdv.var  = "icdv",
    dx.var    = "dx",
    poa       = 1,             # treat all codes as present on admission
    primarydx = 0,             # exclude primary diagnosis codes
    method    = "elixhauser_ahrq2026"
  )

head(elix_sample[, c("patid", "cmrb_flag", "mortality_index", "readmission_index")])
```

### Charlson (Quan 2005)

```{r charlson}
charlson_sample <-
  comorbidities(
    data      = mdcr,
    id.vars   = "patid",
    icd.codes = "code",
    icdv.var  = "icdv",
    dx.var    = "dx",
    poa       = 0,             # POA not required for Charlson
    primarydx = 0,
    method    = "charlson_quan2005"
  )

head(charlson_sample[, c("patid", "cmrb_flag", "cci")])
```

### Pediatric Complex Chronic Conditions (PCCC v3.1)

Flag PCCC conditions and subconditions.

```{r pccc}
pccc_sample <-
  comorbidities(
    data          = mdcr,
    id.vars       = "patid",
    icd.codes     = "code",
    dx.var        = "dx",
    icdv.var      = "icdv",
    poa           = 1,
    method        = "pccc_v3.1",
    subconditions = TRUE
  )

names(pccc_sample)[1:10]
```

## Longitudinal flagging

Use `flag.method = "cumulative"` with at least two `id.vars` (e.g., patient + encounter). The first occurrence of a condition is carried forward to later encounters.

```{r cumulative}
data(mdcr_longitudinal, package = "medicalcoder")

cum_args <- list(
  data        = mdcr_longitudinal,
  id.vars     = c("patid", "enc"),
  icd.codes   = "code",
  icdv.var    = "icdv",
  dx.var      = "dx",
  primarydx   = 0,
  method      = "charlson_quan2011",
  flag.method = "cumulative"
)

charlson_cum <- do.call(comorbidities, c(cum_args, list(poa = 1)))

charlson_cum[charlson_cum$patid == charlson_cum$patid[1], c("patid", "enc", "cmrb_flag")]
```

## What next?

* For method-specific details, see the vignettes: `charlson`, `elixhauser`, `pccc`, and `comorbidities`.
* For ICD utilities, see the `icd` vignette.
* For performance tips, see the benchmarking directory in the repo.
