<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r "setup", include = FALSE}
library(rcmdcheck)
library(knitr)
library(data.table)
options(knitr.kable.NA = '')
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  results = "hide"
)
```

# Testing _medicalcoder_

Along with GitHub Actions and local tests, the workflow in this directory will
test a recent local build of _medicalcoder_ against every major and minor
release of R from 3.5.0 through the latest version, with and without suggested
packages. The tests are run in [Docker](https://www.docker.com/) images based
on the [R-base](https://hub.docker.com/_/r-base) images.

## System Requirements:
To run the tests you need

* [Docker Desktop](https://www.docker.com/products/docker-desktop/)
* [GNU Make](https://www.gnu.org/software/make/)

Just run `make` from this directory.

**NOTE:** When something goes wrong and you need to dig into a specific image
run from this directory.

    docker run -v .:/work/ -it <image>

# Last Testing Results

```{r}
check_logs <- list.files(path = ".", pattern = "00check.log", full.name = TRUE, recursive = TRUE)
checks <- sapply(check_logs, parse_check, simplify = FALSE)

for (i in seq_along(checks)) {
  checks[[i]][["errors"]] <- gsub("\\n", " ", checks[[i]][["errors"]])
  checks[[i]][["warnings"]] <- gsub("\\n", " ", checks[[i]][["warnings"]])
  checks[[i]][["notes"]] <- gsub("\\n", " ", checks[[i]][["notes"]])
}

errors <- lapply(checks, getElement, "errors")
unique(errors)

warnings <- lapply(checks, getElement, "warnings")
unique(warnings)

notes <- lapply(checks, getElement, "notes")
unique(notes)

DT <-
  checks |>
  lapply(
    function(x) {
      with(x,
        data.table(
          rversion,
          status,
          errors   = ifelse(length(errors) > 0, errors, NA_character_),
          warnings = ifelse(length(warnings) > 0, warnings, NA_character_),
          notes    = ifelse(length(notes) > 0, notes, NA_character_),
          checkdir,
          version
        )
      )
    }
  ) |>
  rbindlist()

DT[, suggests := fifelse(grepl("R-.+without-suggests.+$", DT$checkdir), "Without Suggested Packages", "With Suggested Packages")]
DT[, checkdir := NULL]
DT <- unique(DT)

DT[, errors := data.table::fifelse(is.na(errors), "None", errors)]
DT[, warnings := data.table::fifelse(is.na(warnings), "None", warnings)]
DT[, notes := data.table::fifelse(is.na(notes), "None", notes)]

DT[, errors   := factor(errors)]
DT[, warnings := factor(warnings)]
DT[, notes    := factor(notes)]
setkey(DT, suggests, rversion)
```

The green tiles indicate no notes, no warning, no error.

```{r testing-matrix, echo = FALSE, fig.width=10, fig.height=5}
data.table::melt(
  data = DT,
  id.vars = c("rversion", "version", "suggests", "status")
  )[,
    fill := data.table::fcase(
      value == "None", "darkgreen",
      variable == "errors", "darkred",
      variable == "warnings", "red",
      variable == "notes", "yellow"
      )][,
    rv := paste("R version:", sub("\\d$", "x", rversion))
    ][
    , variable := factor(variable, levels = rev(levels(variable)))
    ]|>
ggplot2::ggplot(data = _) +
  ggplot2::theme_bw() +
  ggplot2::aes(
    x = rversion,
    y = variable,
    fill = fill,
  ) +
  ggplot2::geom_tile(color = "white") +
  ggplot2::geom_point(data = function(x) subset(x, value != "None"), mapping = ggplot2::aes(shape = factor(value))) +
  ggplot2::scale_fill_identity() +
  ggplot2::facet_grid(suggests ~ rv, scales = "free_x") +
  ggplot2::guides(shape = ggplot2::guide_legend(ncol = 1)) +
  ggplot2::theme(
    axis.title = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_text(angle = 30, vjust = 1, hjust = 1),
    legend.title = ggplot2::element_blank(),
    legend.position = "bottom"
  )
```
