<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, label = "setup", include = FALSE, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
library(kableExtra)
suppressPackageStartupMessages(library(data.table))
```

# Benchmarking `medicalcoder`

The major factors impacting the expected computation time for applying a
comorbidity algorithm to a data set are:

1. Data size: number of subjects and encounters.
2. Data storage class: `medicalcoder` has been built such that no imports of
   other namespaces is required. That said, when a `data.table` is passed to
   `comorbidities()` and the `data.table` namespace is available, then S3
   dispatch for `merge` is used, along with some other methods, to reduce memory
   use and reduce computation time. When a `tibble` is passed and tidyverse
   namespaces are available, the tibble-aware path reduces time relative to a
   base `data.frame`, though `data.table` remains the fastest option.
3. flag.method: "current" will take less time than the "cumulative" method.

<img src="benchmark2-composite.svg"/>

```{r outtable, include = FALSE}
bench <- readRDS("bench.rds")
setDT(bench)

relative <- readRDS("relative.rds")
setDT(relative)

bench <-
  merge(
    x = bench,
    y = relative,
    by = intersect(names(bench), names(relative))
  )

bench[, memory_gib := max_rss_kib / (1024^2)] # From KiB to GiB


outtable <-
  dcast(
    data = bench,
    data_class + subjects + method + subconditions ~ flag.method,
    value.var = list("encounters", "time_seconds", "relative_time", "memory_gib"),
    fun.aggregate = list(mean, mean, mean, mean)
  )
set(outtable, j = "encounters_mean_cumulative", value = NULL)
setnames(outtable, old = "encounters_mean_current", new = "encounters_mean")
outtable_orig <- data.table::copy(outtable)

neworder <-
  c("method",
    "subconditions",
    "subjects",
    "encounters_mean",
    "data_class",
    "time_seconds_mean_current",
    "relative_time_mean_current",
    "memory_gib_mean_current",
    "time_seconds_mean_cumulative",
    "relative_time_mean_cumulative",
    "memory_gib_mean_cumulative"
  )

setcolorder(outtable, neworder = neworder)

outtable[, data_class := factor(data_class, levels = c("data.frame", "data.table", "tibble"))]
setkey(outtable, method, subconditions, subjects, data_class)

outtable[, subjects := qwraps2::frmt(as.integer(subjects))]

outtable[, method := fifelse(subconditions, paste(method, "(with subconditions)"), method)]
outtable[, subconditions := NULL]
```

In general, the expected time to apply a comorbidity method is lower for
`tibble`s than for base `data.frame`s, and lower still for `data.table`s. Best
observed case: a `data.table` took
`r outtable_orig[data_class == "data.table", min(relative_time_mean_current, relative_time_mean_cumulative)]`
the time of a `data.frame`. Best case for `tibble`s was
`r outtable_orig[data_class == "tibble", min(relative_time_mean_current, relative_time_mean_cumulative)]`
the time of a `data.frame`.

```{r show-outtable, echo = FALSE, results = "asis"}
foo <- function(x) {
  kbl(
    x = x,
    format = "html",
    escape = FALSE,
    digits = 2,
    align = "rrrrrrrrr",
    col.names = c("Subjects", "Encounters", "Data Class", rep(c("Time (seconds)", "Relative time", "Memory (GB)"), times = 2)),
    caption = "Expected time (seconds), relative time (with respect to data.frame), and expected memory use, by flagging method (current or cumulative), number of subjects, average number of total encounters, and input data storage format."
  ) |>
  collapse_rows(columns = 1, valign = "middle") |>
  row_spec(row = c(seq(1, nrow(x), by = 6)), background = "#D9D9D9") |>
  row_spec(row = c(seq(2, nrow(x), by = 6)), background = "#D9D9D9") |>
  row_spec(row = c(seq(3, nrow(x), by = 6)), background = "#D9D9D9") |>
  add_header_above(c("", "", "", "flag.method = 'current'" = 3, "flag.method = 'cumulative'" = 3))
}

cat("\n### Benchmarking Charlson (Quan 2005)\n\n")
foo(x = outtable[method == "charlson_quan2005", -"method"])

cat("\n### Benchmarking Elixhauser (Quan 2005)\n\n")
foo(x = outtable[method == "elixhauser_quan2005", -"method"])

cat("\n### Benchmarking PCCC v3.1 (without subconditions)\n\n")
foo(x = outtable[method == "pccc_v3.1", -"method"])

cat("\n### Benchmarking PCCC v3.1 (with subconditions)\n\n")
foo(x = outtable[method == "pccc_v3.1 (with subconditions)", -"method"])
```
