<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, label = "setup", include = FALSE, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
library(knitr)
library(kableExtra)
suppressPackageStartupMessages(library(data.table))
```

# Benchmarking `medicalcoder`

The major factors impacting the expected computation time for applying a
comorbidity algorithm to a data set are:

1. Data size: number of encounters.
2. Data storage class: `medicalcoder` has been built such that no imports of
   other namespaces is required. That said, when a `data.table` is passed to
   `comorbidities()` and the `data.table` namespace is available, then S3
   dispatch for `merge` is used, along with some other methods, to reduce memory
   use and reduce computation time. When a `tibble` is passed and tidyverse
   namespaces are available, the tibble-aware path reduces time relative to a
   base `data.frame`, though `data.table` remains the fastest option.
3. flag.method: "current" will take less time than the "cumulative" method.

In the following examples, each generated data set was benchmark by

* `method`:
    * charlson_quan2005
    * elixhauser_quan2005
    * pccc_v3.1 without subconditions
    * pccc_v3.1 with subconditions
* data class:
    * data.frame
    * tibble
    * data.table
* `flag.method`
    * current
    * cumulative

```{r data-import, include = FALSE}
mimicivdata <-
  data.table::as.data.table(
    readRDS(file.path("mimic-iv", "mimicivdata.rds"))
  )
mimicivbenchmark <-
  data.table::as.data.table(
    readRDS(file.path("mimic-iv", "benchmark.rds"))
  )
```

```{r mimic-pccc-v3.1, include = FALSE}
mimicivpcccv3 <- medicalcoder::comorbidities(
  data = mimicivdata,
  id.vars = c("subject_id"),
  icd.codes = "icd_code",
  dx.var = "dx",
  poa = 1,
  flag.method = "current",
  method = "pccc_v3.1"
)
```

```{r mimic-charlson, include = FALSE}
mimicivcharlson <- medicalcoder::comorbidities(
  data = mimicivdata,
  id.vars = c("subject_id"),
  icd.codes = "icd_code",
  dx.var = "dx",
  poa = 1,
  flag.method = "current",
  method = "charlson_quan2005",
  primarydx = 0
)
```

```{r mimic-elixhauser, include = FALSE}
mimicivelixhauser <- medicalcoder::comorbidities(
  data = mimicivdata,
  id.vars = c("subject_id"),
  icd.codes = "icd_code",
  dx.var = "dx",
  poa = 1,
  flag.method = "current",
  method = "elixhauser_quan2005",
  primarydx = 0
)
```


## MIMIC-IV Demo Data

The following summary is based on resampling the [MIMIC-IV Clinical Database
Demo](https://physionet.org/content/mimic-iv-demo/2.2/) data set provided by
PhysioNet.[^1][^2] Using the MIMIC-IV Clinical Database Demo set will allow others to benchmark
`medicalcoder` on their system with the same data as used here.

[^1]: Johnson, A., Bulgarelli, L., Pollard, T., Horng, S., Celi, L. A., & Mark, R. (2023). MIMIC-IV Clinical Database Demo (version 2.2). PhysioNet. RRID:SCR_007345. https://doi.org/10.13026/dp1f-ex47

[^2]: Goldberger, A., Amaral, L., Glass, L., Hausdorff, J., Ivanov, P. C., Mark, R., ... & Stanley, H. E. (2000). PhysioBank, PhysioToolkit, and PhysioNet: Components of a new research resource for complex physiologic signals. Circulation [Online]. 101 (23), pp. e215–e220. RRID:SCR_007345.

The MIMIC-IV demo data consists of
`r data.table::uniqueN(mimicivdata, by = c("subject_id"))` subjects and a total of
`r data.table::uniqueN(mimicivdata, by = c("subject_id", "hadm_id"))`
encounters.  A mix of ICD-9 and ICD-10 diagnostic and procedure codes are in the
data.

`r qwraps2::n_perc(mimicivpcccv3[["cmrb_flag"]])` of the subjects had at least
one PCCC comorbidity,
`r qwraps2::n_perc(mimicivcharlson[["cmrb_flag"]])`
of the subjects had at least one Charlson comorbidity, and
`r qwraps2::n_perc(mimicivelixhauser[["cmrb_flag"]])`
of the subjects had at least one Elixhauser comorbidity.

The subjects were resampled to generate data sets with upto 1,000,000 subjects.

<img src="benchmark_mimiciv_composite.svg"/>

In general, the expected time to apply a comorbidity method is lower for
`tibble`s than for base `data.frame`s, and lower still for `data.table`s. Best
observed case: a `data.table` took
`r mimicivbenchmark[data_class == "data.table", min(relative_time)]`
the time of a `data.frame`. Best case for `tibble`s was
`r mimicivbenchmark[data_class == "tibble", min(relative_time)]`
the time of a `data.frame`.

## All Comorbidities

Some of the computational expense in building the indicators for the
comorbidities is due to the number of subjects/encounters with at least one
comorbidity. We generated data sets where all subjects on all encounters had at
least one ICD code mapping to a comorbidity.  The benchmarking results are
summarized in the following graphic.

<img src="benchmark_all_cmrb_composite.svg"/>
