---
title: "medicalcoder vs comorbidity"
output:
 rmarkdown::html_vignette:
   toc: true
   number_sections: false
bibliography: ../references.bib
vignette: >
 %\VignetteIndexEntry{medicalcoder vs comorbidity}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r, label = "setup", include = FALSE}
# IMPORTANT SYNTAX NOTE:
# Vignettes cannot use the pipeOp `|>` so the package can depend on R >= 3.5.0
# The _articles_ (website only) can use the pipeOp as since these are not part of the source package.

library(kableExtra)
options(qwraps2_markup = "markdown")
options(knitr.kable.NA = '')
knitr::opts_chunk$set(collapse = TRUE, fig.align = "center")
```

# Introduction

The purpose of this article is to compare the API and results between
`medicalcoder` and the R package
[comorbidity](https://cran.r-project.org/package=comorbidity)
[@r-comorbidity].

```{r}
library(medicalcoder)
library(comorbidity)
packageVersion("comorbidity")
packageDescription("comorbidity")$Title
cat(packageDescription("comorbidity")$Description)
```

The following table is a partical list of similarities and differences between
`medicalcoder` and `comorbidity`.

```{r, echo = FALSE, results = "asis"}
x <- data.table::fread(text ="
feature | `comorbidity` | `medicalcoder`
Charlson [@quan2005]                     | &#9989;       | &#9989;
Charlson [@quan2011]                     | &#10060;      | &#9989;
Charlson [@deyo1992adapting]             | &#10060;      | &#9989;
Charlson [@glasheen2019]                 | &#10060;      | &#9989;
Charlson (Ludvigsson et al., 2021)       | &#9989;       | &#10060;
Charlson (Sundararajan et al., 2004)     | &#9989;       | &#10060;
Elixhauser [@quan2005]                   | &#9989;       | &#9989;
Elixhauser [@elixhauser1998]             | &#10060;      | &#9989;
Elixhauser (ICD-9-CM AHRQ)               | &#10060;      | &#9989;
Elixhauser (ICD-10-CM AHRQ)              | &#10060;      | &#9989;
PCCC version 2 [@feudtner2014pediatric]  | &#10060;      | &#9989;
PCCC version 3 [@feinstein2024pediatric] | &#10060;      | &#9989;
Allows for full ICD codes | &#10060; | &#9989;
Allows for compact ICD codes | &#9989; | &#9989;
Applys algorithms to ICD-9 and ICD-10 simulataneously | &#10060 | &#9989;
Diagnosis and Procedure codes simulataneously | <sup>*</sup> | &#9989;
Look up tables <sup>**</sup>              | &#9989;       | &#9989;
Prcoedure codes                          | &#10060;      | &#9989;
Disagnostic codes                        | &#9989;       | &#9989;
Functions for looking up/testing ICD Codes | &#10060;    | &#9989;
")

y <- data.table::as.data.table(get_icd_codes())[
  , .(min(known_start), max(known_end), length(unique(c(known_start, known_end))))
  , keyby = .(icdv, factor(dx, 0:1, c("Procedure", "Diagnosis")), toupper(src))]
y <-
kbl(
  y,
  col.names = c("Version", "Type", "Source", "First", "Last", "Distinct"),
  align = "cccccc"
  ) |>
  kable_styling(bootstrap_options = c("striped"), font_size = 8) |>
  add_header_above(c("ICD" = 2, "", "Years" = 3)) |>
  add_header_above(c("`medicalcoder` known ICD codes" = 6)) |>
  as.character()

kbl(
  x,
  escape = FALSE,
  col.names = c("", "`comorbidity`", "`medicalcoder`"),
  align = "lcc"
) |>
kable_styling(bootstrap_options = c("striped")) |>
pack_rows("Comorbidity Algorithm and/or Scoring", start_row = 1, end_row = 12) |>
pack_rows("Comorbidity ICD Code Inputs", start_row = 13, end_row = 16) |>
pack_rows("ICD Utilities", start_row = 17, end_row = nrow(x)) |>
footnote(symbol_manual = c("*", "**"),
  symbol = c("Procedure codes are not relevent to Charlson and Elixhauser, not
    needed in `comorbidity`.",
  paste(
"<small>`comorbidity` contains look up tables for ICD-10 years 2009, 2011,
  ICD-10-CM 2018, 2022, and ICD-9 for 2015.  `medicalcoder` has a more
  extensive internal ICD code database.</small>", y)), escape = FALSE)
```

# Prepare Data for `comorbidity`
The example data set `mdcr` within `medicalcoder` is in a format that is ideal
for `medicalcoder`.  To prepare that data set for use in the
`comorbidity::comorbidity()` call we need to split the data in ICD-9 and ICD-10
version. Also, `medicalcoder` can handle diagnostic and procedure codes
simulataneously; a nessecary feature for PCCC.  `comorbidity` only applies
comorbidity algorithms based on diagnostic codes and thus does not expect to see
any procedure codes.

```{r}
mdcr_icd9dx  <- subset(mdcr, icdv ==  9L & dx == 1L)
mdcr_icd10dx <- subset(mdcr, icdv == 10L & dx == 1L)
```
To make sure we have a fair comparison, we will add rows without a code for each
patid not in the subsets.
```{r}
mdcr_icd9dx <-
  rbind(
    mdcr_icd9dx,
    data.frame(patid = setdiff(mdcr$patid, mdcr_icd9dx$patid), icdv = 9L, code = "", dx = 1L)
  )
mdcr_icd10dx <-
  rbind(
    mdcr_icd10dx,
    data.frame(patid = setdiff(mdcr$patid, mdcr_icd10dx$patid), icdv = 10L, code = "", dx = 1L)
  )
```

# Charlson Comorbidities

Appling the Charlson comorbidities to the `mdcr` data set via `medicalcoder` is
done as follows.  Imporant note: the ICD codes used to assess the Charlson
comorbidities between @quan2005 and @quan2011 are the same. The scoring is not.
To match the scoring used in `comorbidity` we need to use the
`charlson_quan2011` method in `medicalcoder`.

```{r}
medicalcoder_charlson_results <-
  medicalcoder::comorbidities(
    data = mdcr,
    id.vars = "patid",
    icd.codes = "code",
    dx.var = "dx",
    icdv.var = "icdv",
    poa = 1L, # assume all codes are present on admission
    primarydx = 0L, # assume all codes are secondary diagnoses
    method = "charlson_quan2011"
  )
```

For `comorbidity`, we will need to make two calls, one for the ICD-9 codes and
one for the ICD-10 codes.  We will then need to combine the results and then
apply a scoring function.

```{r}
comorbidity_charlson_icd9_results <-
  comorbidity::comorbidity(
    x = mdcr_icd9dx,
    id = "patid",
    code = "code",
    map = "charlson_icd9_quan",
    assign0 = TRUE # set less severe comorbidities flags to 0 when more severe comorbidities is also flagged
  )

comorbidity_charlson_icd10_results <-
  comorbidity::comorbidity(
    x = mdcr_icd10dx,
    id = "patid",
    code = "code",
    map = "charlson_icd10_quan",
    assign0 = TRUE
  )

# combine the ICD-9 and ICD-10 results into one set
comorbidity_charlson_results <-
  rbind(comorbidity_charlson_icd9_results, comorbidity_charlson_icd10_results)

comorbidity_charlson_results <-
  aggregate(. ~ patid, data = comorbidity_charlson_results, FUN = max)

# add the attributes to the combine set
attributes(comorbidity_charlson_results)[c("class", "variable.labels", "map")] <-
  attributes(comorbidity_charlson_icd9_results)[c("class", "variable.labels", "map")]

comorbidity_charlson_results[["score"]] <-
  comorbidity::score(
    x = comorbidity_charlson_results,
    weights = "quan",
    assign0 = TRUE
  )

# the score return is a numeric value, setting to integer to match the storage
# mode of medicalcoder
comorbidity_charlson_results[["score"]] <-
  as.integer(comorbidity_charlson_results[["score"]])
```

We can now compare the comorbidity flags between the two packages.  First we
check that we have the same number of rows and all the patid are accounted for.

```{r}
nrow(medicalcoder_charlson_results) == nrow(comorbidity_charlson_results)

all(mdcr$patid %in% medicalcoder_charlson_results$patid)
all(medicalcoder_charlson_results$patid %in% mdcr$patid)

all(mdcr$patid %in% comorbidity_charlson_results$patid)
all(comorbidity_charlson_results$patid %in% mdcr$patid)
```

There are differences in the results in terms of the names and numbers of
columns returned.

```{r}
dim(medicalcoder_charlson_results)
names(medicalcoder_charlson_results)

dim(comorbidity_charlson_results)
names(comorbidity_charlson_results)
```

We will compare results column by column.  First, we merge the data sets
together by patid.

```{r}
charlson_delta <-
  merge(
    x = comorbidity_charlson_results,
    y = medicalcoder_charlson_results,
    all = TRUE,
    by = "patid"
  )
```

The relevent columns:

```{r, include = FALSE}
charlson_columns <- data.table::fread(text = "
Condition                          | comorbidity | medicalcoder
AIDS                               | aids        | aidshiv
Cancer (Any malignancy)            | canc        | mal
Cancer (Metastatic solid tumor)    | metacanc    | mst
Cerebrovascular disease            | cevd        | cebvd
Chronic pulmonary disease          | cpd         | copd
Congestive heart failure           | chf         | chf
Dementia                           | dementia    | dem
Hemiplegia or paraplegia           | hp          | hp
Myocardial infarction              | mi          | mi
Peripheral vascular disease        | pvd         | pvd
Peptic ulcer disease               | pud         | pud
Rheumatic disease                  | rheumd      | rhd
Diabetes (uncomplicated)           | diab        | dm
Diabetes (complicated)             | diabwc      | dmc
Renal disease                      | rend        | rnd
Liver disease (mild)               | mld         | mld
Liver disease (moderate or severe) | msld        | msld
Score (based on Quan 2011)         | score       | cci
  ")
data.table::setDF(charlson_columns)
```

The following data.frame reports the condition and the corresponding column from
`comorbidity` and from `medicalcoder`.
```{r}
charlson_columns
```

For each row in `charlson_columns` we will verify that the results are
identical in `charlson_delta` and remove the columns.

```{r, echo = TRUE, results = "markup"}
for (i in seq_len(nrow(charlson_columns))) {
  x <- charlson_columns[["comorbidity"]][i]
  y <- charlson_columns[["medicalcoder"]][i]
  if (x == y) {
    x <- paste0(x, ".x")
    y <- paste0(y, ".y")
  }
  e <- base::substitute(identical(charlson_delta[[X]], charlson_delta[[Y]]), list(X = x, Y = y))
  print(e)
  r <- eval(e)
  print(r)
  stopifnot(r)
  charlson_delta[[x]] <- NULL
  charlson_delta[[y]] <- NULL
}
```

The only columns left in the `charlson_delta` object are from `medicalcoder`

```{r}
str(charlson_delta)
```

* `num_cmrb`: number of comorbidities flagged
* `cmrb_flag`: a 0/1 indicator for any comorbidity
* `age_score`: a score adjustment based on age _if_ age is provided.

```{r}
stopifnot(
  identical(
    names(charlson_delta),
    c("patid", "num_cmrb", "cmrb_flag", "age_score")
  )
)
```

An addition feature `medicalcoder` provides that `comorbidity` does not is a
summary method for the results, which can be used to build nice tables via
[kableExtra](https://CRAN.R-project.org/package=kableExtra).

```{r}
x <- summary(medicalcoder_charlson_results)[[1]][, c("condition_description", "count", "percent")]

kbl(
  x = x,
  digits = 2,
  col.names = c("", "Count", "Percentage")
) |>
pack_rows("Comorbidity", start_row = 1, end_row = 17) |>
pack_rows("Number of Comorbidities", start_row = 18, end_row = nrow(x))
```

# Elixhauser Comorbidities

Appling the Elixhauser comorbidities to the `mdcr` data set via `medicalcoder` is
done as follows.

```{r}
medicalcoder_elixhauser_results <-
  medicalcoder::comorbidities(
    data = mdcr,
    id.vars = "patid",
    icd.codes = "code",
    dx.var = "dx",
    icdv.var = "icdv",
    poa = 1L,       # assume all codes are present on admission
    primarydx = 0L, # assume all codes are secondary diagnoses
    method = "elixhauser_quan2005"
  )
```

For `comorbidity`, we will need to make two calls, one for the ICD-9 codes and
one for the ICD-10 codes.  We will then need to combine the results and then
apply a scoring function.

```{r}
comorbidity_elixhauser_icd9_results <-
  comorbidity::comorbidity(
    x = mdcr_icd9dx,
    id = "patid",
    code = "code",
    map = "elixhauser_icd9_quan",
    assign0 = TRUE # set less severe comorbidities flags to 0 when more severe comorbidities is also flagged
  )

comorbidity_elixhauser_icd10_results <-
  comorbidity::comorbidity(
    x = mdcr_icd10dx,
    id = "patid",
    code = "code",
    map = "elixhauser_icd10_quan",
    assign0 = TRUE
  )

# combine the ICD-9 and ICD-10 results into one set
comorbidity_elixhauser_results <-
  rbind(comorbidity_elixhauser_icd9_results, comorbidity_elixhauser_icd10_results)

comorbidity_elixhauser_results <-
  aggregate(. ~ patid, data = comorbidity_elixhauser_results, FUN = max)
```

We can now compare the comorbidity flags between the two packages.  First we
check that we have the same number of rows and all the patid are accounted for.

```{r}
nrow(medicalcoder_elixhauser_results) == nrow(comorbidity_elixhauser_results)

all(mdcr$patid %in% medicalcoder_elixhauser_results$patid)
all(medicalcoder_elixhauser_results$patid %in% mdcr$patid)

all(mdcr$patid %in% comorbidity_elixhauser_results$patid)
all(comorbidity_elixhauser_results$patid %in% mdcr$patid)
```

There are differences in the results in terms of the names and numbers of
columns returned.

```{r}
dim(medicalcoder_elixhauser_results)
names(medicalcoder_elixhauser_results)

dim(comorbidity_elixhauser_results)
names(comorbidity_elixhauser_results)
```

We will compare results column by column.  First, we merge the data sets
together by patid.

```{r}
elixhauser_delta <-
  merge(
    x = comorbidity_elixhauser_results,
    y = medicalcoder_elixhauser_results,
    all = TRUE,
    by = "patid"
  )
```

The relevent columns:

```{r, include = FALSE}
#Hypertension (Uncomplicated or Complicated)   |             | HTN_C
elixhauser_columns <- data.table::fread(text = "
Condition                                     | comorbidity | medicalcoder
AIDS/HIV                                      | aids        | AIDS
Alcohol abuse                                 | alcohol     | ALCOHOL
Anemias (Blood loss anaemia)                  | blane       | BLDLOSS
Anemias (Deficiency anaemia)                  | dane        | ANEMDEF
Cardiac Arrhythmias                           | carit       | CARDIAC_ARRHYTHMIAS
Cancer (Solid tummor without metastasis)      | solidtum    | TUMOR
Cancer (Metastatic)                           | metacanc    | METS
Chronic pulmonary disease                     | cpd         | CHRNLUNG
Coagulopathy                                  | coag        | COAG
Congestive Heart Failure                      | chf         | CHF
Depression                                    | depre       | DEPRESS
Diabetes (Complicated)                        | diabc       | DMCX
Diabetes (Uncomplicated)                      | diabunc     | DM
Drug abuse                                    | drug        | DRUG
Fluid and electrolyte disorders               | fed         | LYTES
Hypertension (Complicated)                    | hypc        | HTN_CX
Hypertension (Uncomplicated)                  | hypunc      | HTN_UNCX
Hypothyroidism                                | hypothy     | HYPOTHY
Liver disease                                 | ld          | LIVER
Lymphoma                                      | lymph       | LYMPH
Obesity                                       | obes        | OBESE
Other neurological disease                    | ond         | NEURO
Paralysis                                     | para        | PARA
Peptic ulcer disease excluding bleeding       | pud         | ULCER
Peripheral Vascular Disorders                 | pvd         | PERIVASC
Psychoses                                     | psycho      | PSYCH
Pulmonary Circulation Disorders               | pcd         | PULMCIRC
Renal Failure                                 | rf          | RENLFAIL
Rheumatoid artritis/collaged vascular disease | rheumd      | ARTH
Valvular disease                              | valv        | VALVE
Weight loss                                   | wloss       | WGHTLOSS
  ")
data.table::setDF(elixhauser_columns)
```

The following data.frame reports the condition and the corresponding column from
`comorbidity` and from `medicalcoder`.
```{r}
elixhauser_columns
```

For each row in `elixhauser_columns` we will verify that the results are
identical in `elixhauser_delta` and remove the columns.


```{r, include = FALSE}
# the HTN_C is only in the medicalcoder results
stopifnot(
  identical(
    medicalcoder_elixhauser_results[["HTN_C"]],
    as.integer(
      medicalcoder_elixhauser_results[["HTN_UNCX"]] |
        medicalcoder_elixhauser_results[["HTN_CX"]]
    )
  )
)
```

```{r, echo = TRUE, results = "markup"}
for (i in seq_len(nrow(elixhauser_columns))) {
  x <- elixhauser_columns[["comorbidity"]][i]
  y <- elixhauser_columns[["medicalcoder"]][i]
  if (x == y) {
    x <- paste0(x, ".x")
    y <- paste0(y, ".y")
  }
  e <- base::substitute(identical(elixhauser_delta[[X]], elixhauser_delta[[Y]]), list(X = x, Y = y))
  print(e)
  r <- eval(e)
  print(r)
  stopifnot(r)
  elixhauser_delta[[x]] <- NULL
  elixhauser_delta[[y]] <- NULL
}
```

The remaining columns in the `elixhauser_delta` object are specific to
`medicalcoder`, save patid.

```{r}
str(elixhauser_delta)
```

```{r, inlcude = FALSE}
stopifnot(identical(names(elixhauser_delta), c("patid", "HTN_C", "num_cmrb", "cmrb_flag", "mortality_index", "readmission_index")))
```

* `HTN_C`: _any_ hypertension - added to simplify the cacluation of the
  mortality and readmission indices based on weights from AHRQ.

* `num_cmrb`:  Number of comorbidities flagged

* `cmrb_flag`: a 0/1 indicator for _any_ comorbidity being flagged

* `mortality_index` and `readmission_index`: index scores based on the AHRQ
  weights [@ahrq2017].

An addition feature `medicalcoder` provides that `comorbidity` does not is a
summary method for the results, which can be used to build nice tables via
[kableExtra](https://CRAN.R-project.org/package=kableExtra).

```{r}
x <- summary(medicalcoder_elixhauser_results)[[1]][, c("condition", "count", "percent")]

kbl(
  x = x,
  digits = 2,
  col.names = c("", "Count", "Percentage")
) |>
pack_rows("Comorbidity", start_row = 1, end_row = which(x$condition == ">= 1") - 1L) |>
pack_rows("Number of Comorbidities", start_row = which(x$condition == ">= 1"), end_row = nrow(x))
```

# References

<!-- ----------------------------------------------------------------------- -->
<!--                              End of File                                -->
<!-- ----------------------------------------------------------------------- -->
