---
title: "medicalcoder vs comorbidity"
output:
 rmarkdown::html_vignette:
   toc: true
   number_sections: false
bibliography: ../references.bib
vignette: >
 %\VignetteIndexEntry{medicalcoder vs comorbidity}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r, label = "setup", include = FALSE}
# IMPORTANT SYNTAX NOTE:
# Vignettes cannot use the pipeOp `|>` so the package can depend on R >= 3.5.0
# The _articles_ (website only) can use the pipeOp since these are not part of the source package.

library(kableExtra)
options(qwraps2_markup = "markdown")
options(knitr.kable.NA = '')
knitr::opts_chunk$set(collapse = TRUE, fig.align = "center")
```

# Introduction

The purpose of this article is to compare the API and results between
medicalcoder and the R package
[comorbidity](https://cran.r-project.org/package=comorbidity)
[@r-comorbidity].

```{r}
library(medicalcoder)
library(comorbidity)
packageVersion("comorbidity")
packageDescription("comorbidity")$Title
cat(packageDescription("comorbidity")$Description)
```

```{r, include = FALSE}
stopifnot(packageVersion("comorbidity") == "1.1.0")
```

The following table is a partial list of similarities and differences between
medicalcoder and comorbidity.

```{r, echo = FALSE, results = "asis"}
x <- data.table::fread(text ="
feature                                               | comorbidity  | medicalcoder
Charlson [@quan2005]                                  | &#9989;      | &#9989;
Charlson [@quan2011]                                  | &#10060;     | &#9989;
Charlson [@deyo1992adapting]                          | &#10060;     | &#9989;
Charlson [@glasheen2019]                              | &#10060;     | &#9989;
Charlson (Ludvigsson et al., 2021)                    | &#9989;      | &#9989;
Charlson (Sundararajan et al., 2004)                  | &#9989;      | &#9989;
Elixhauser [@quan2005]                                | &#9989;      | &#9989;
Elixhauser [@elixhauser1998]                          | &#10060;     | &#9989;
Elixhauser (ICD-9-CM AHRQ)                            | &#10060;     | &#9989;
Elixhauser (ICD-10-CM AHRQ)                           | &#10060;     | &#9989;
PCCC version 2 [@feudtner2014pediatric]               | &#10060;     | &#9989;
PCCC version 3 [@feinstein2024pediatric]              | &#10060;     | &#9989;
Allows for full ICD codes                             | &#10060;     | &#9989;
Allows for compact ICD codes                          | &#9989;      | &#9989;
Applies algorithms to ICD-9 and ICD-10 simultaneously | &#10060      | &#9989;
Diagnosis and Procedure codes simultaneously          | <sup>*</sup> | &#9989;
Lookup tables <sup>**</sup>                           | &#9989;      | &#9989;
Procedure codes                                       | &#10060;     | &#9989;
Diagnostic codes                                      | &#9989;      | &#9989;
Functions for looking up/testing ICD Codes            | &#10060;     | &#9989;
")

y <- data.table::as.data.table(get_icd_codes())[
  , .(min(known_start), max(known_end), length(unique(c(known_start, known_end))))
  , keyby = .(icdv, factor(dx, 0:1, c("Procedure", "Diagnosis")), toupper(src))]
y <-
kbl(
  y,
  col.names = c("Version", "Type", "Source", "First", "Last", "Distinct"),
  align = "cccccc"
  ) |>
  kable_styling(bootstrap_options = c("striped"), font_size = 8) |>
  add_header_above(c("ICD" = 2, "", "Years" = 3)) |>
  add_header_above(c("medicalcoder known ICD codes" = 6)) |>
  as.character()

kbl(
  x,
  escape = FALSE,
  col.names = c("", "comorbidity", "medicalcoder"),
  align = "lcc"
) |>
kable_styling(bootstrap_options = c("striped")) |>
pack_rows("Comorbidity Algorithm and/or Scoring", start_row = 1, end_row = 12) |>
pack_rows("Comorbidity ICD Code Inputs", start_row = 13, end_row = 16) |>
pack_rows("ICD Utilities", start_row = 17, end_row = nrow(x)) |>
footnote(symbol_manual = c("*", "**"),
  symbol = c("Procedure codes are not relevant to Charlson and Elixhauser, not
    needed in comorbidity.",
  paste(
"<small>comorbidity contains look up tables for ICD-10 years 2009, 2011,
  ICD-10-CM 2018, 2022, and ICD-9 for 2015.  medicalcoder has a more
  extensive internal ICD code database.</small>", y)), escape = FALSE)
```

# Prepare Data for comorbidity

The example data set `mdcr` within medicalcoder is in a format that is ideal
for `medicalcoder::comorbidities()`.  To prepare that data set for use in the
`comorbidity::comorbidity()` call we need to split the data in ICD-9 and ICD-10
version. Also, `medicalcoder::comorbidities()` can handle diagnostic and
procedure codes simultaneously; a necessary feature for PCCC.
`comorbidity::comorbidity()` only applies comorbidity algorithms based on
diagnostic codes and thus does not expect to see any procedure codes.

```{r}
mdcr_icd9dx  <- subset(mdcr, icdv ==  9L & dx == 1L)
mdcr_icd10dx <- subset(mdcr, icdv == 10L & dx == 1L)
```
To make sure we have a fair comparison, we will add rows without a code for each
patid not in the subsets.
```{r}
mdcr_icd9dx <-
  rbind(
    mdcr_icd9dx,
    data.frame(patid = setdiff(mdcr$patid, mdcr_icd9dx$patid), icdv = 9L, code = "", dx = 1L)
  )
mdcr_icd10dx <-
  rbind(
    mdcr_icd10dx,
    data.frame(patid = setdiff(mdcr$patid, mdcr_icd10dx$patid), icdv = 10L, code = "", dx = 1L)
  )
```

# Charlson Comorbidities

Applying the Charlson comorbidities to the `mdcr` data set via
`medicalcoder::comorbidities()` is
done as follows.  Important note: the ICD codes used to assess the Charlson
comorbidities between @quan2005 and @quan2011 are the same. The scoring is not.
To match the scoring used in `comorbidity::comorbidity()` we need to use the
`method = "charlson_quan2011"` in `medicalcoder::comorbidities()`.

```{r}
medicalcoder_charlson_results <-
  medicalcoder::comorbidities(
    data = mdcr,
    id.vars = "patid",
    icd.codes = "code",
    dx.var = "dx",
    icdv.var = "icdv",
    poa = 1L, # assume all codes are present on admission
    primarydx = 0L, # assume all codes are secondary diagnoses
    method = "charlson_quan2011"
  )
```
The return from `medicalcoder::comorbidities()` is a list object. We only need
the "conditions" element for this following comparisons.
```{r}
str(medicalcoder_charlson_results, max.level = 1)
```

For `comorbidity::comorbidity()`, we will need to make two calls, one for the
ICD-9 codes and one for the ICD-10 codes.  We will then need to combine the
results and then apply a scoring function.

```{r}
comorbidity_charlson_icd9_results <-
  comorbidity::comorbidity(
    x = mdcr_icd9dx,
    id = "patid",
    code = "code",
    map = "charlson_icd9_quan",
    assign0 = TRUE # set less severe comorbidities flags to 0 when more severe comorbidities is also flagged
  )

comorbidity_charlson_icd10_results <-
  comorbidity::comorbidity(
    x = mdcr_icd10dx,
    id = "patid",
    code = "code",
    map = "charlson_icd10_quan",
    assign0 = TRUE
  )

# combine the ICD-9 and ICD-10 results into one set
comorbidity_charlson_results <-
  rbind(comorbidity_charlson_icd9_results, comorbidity_charlson_icd10_results)

comorbidity_charlson_results <-
  aggregate(. ~ patid, data = comorbidity_charlson_results, FUN = max)

# add the attributes to the combine set
attributes(comorbidity_charlson_results)[c("class", "variable.labels", "map")] <-
  attributes(comorbidity_charlson_icd9_results)[c("class", "variable.labels", "map")]

comorbidity_charlson_results[["score"]] <-
  comorbidity::score(
    x = comorbidity_charlson_results,
    weights = "quan",
    assign0 = TRUE
  )

# the score return is a numeric value, setting to integer to match the storage
# mode of medicalcoder
comorbidity_charlson_results[["score"]] <-
  as.integer(comorbidity_charlson_results[["score"]])
```

We can now compare the comorbidity flags between the two packages.  First we
check that we have the same number of rows and all the patid are accounted for.

```{r}
nrow(medicalcoder_charlson_results[["conditions"]]) == nrow(comorbidity_charlson_results)

all(mdcr$patid %in% medicalcoder_charlson_results[["conditions"]][["patid"]])
all(medicalcoder_charlson_results[["conditions"]][["patid"]] %in% mdcr$patid)

all(mdcr$patid %in% comorbidity_charlson_results$patid)
all(comorbidity_charlson_results$patid %in% mdcr$patid)
```

There are differences in the results in terms of the names and numbers of
columns returned.

```{r}
dim(medicalcoder_charlson_results[["conditions"]])
names(medicalcoder_charlson_results[["conditions"]])

dim(comorbidity_charlson_results)
names(comorbidity_charlson_results)
```

We will compare results column by column.  First, we merge the data sets
together by patid.

```{r}
charlson_delta <-
  merge(
    x = comorbidity_charlson_results,
    y = medicalcoder_charlson_results[["conditions"]],
    all = TRUE,
    by = "patid"
  )
```

The relevant columns:

```{r, include = FALSE}
charlson_columns <- data.table::fread(text = "
Condition                          | comorbidity | medicalcoder
AIDS                               | aids        | aidshiv
Cancer (Any malignancy)            | canc        | mal
Cancer (Metastatic solid tumor)    | metacanc    | mst
Cerebrovascular disease            | cevd        | cebvd
Chronic pulmonary disease          | cpd         | copd
Congestive heart failure           | chf         | chf
Dementia                           | dementia    | dem
Hemiplegia or paraplegia           | hp          | hp
Myocardial infarction              | mi          | mi
Peripheral vascular disease        | pvd         | pvd
Peptic ulcer disease               | pud         | pud
Rheumatic disease                  | rheumd      | rhd
Diabetes (uncomplicated)           | diab        | dm
Diabetes (complicated)             | diabwc      | dmc
Renal disease                      | rend        | rnd
Liver disease (mild)               | mld         | mld
Liver disease (moderate or severe) | msld        | msld
Score (based on Quan 2011)         | score       | cci
  ")
data.table::setDF(charlson_columns)
```

The following data.frame reports the condition and the corresponding column from
`comorbidity::comorbidity()` and from `medicalcoder::comorbidities()`.
```{r}
charlson_columns
```

For each row in `charlson_columns` we will verify that the results are
identical in `charlson_delta` and remove the columns.

```{r, echo = TRUE, results = "markup"}
for (i in seq_len(nrow(charlson_columns))) {
  x <- charlson_columns[["comorbidity"]][i]
  y <- charlson_columns[["medicalcoder"]][i]
  if (x == y) {
    x <- paste0(x, ".x")
    y <- paste0(y, ".y")
  }
  e <- base::substitute(identical(charlson_delta[[X]], charlson_delta[[Y]]), list(X = x, Y = y))
  print(e)
  r <- eval(e)
  print(r)
  #stopifnot(r)
  if (r) {
    charlson_delta[[x]] <- NULL
    charlson_delta[[y]] <- NULL
  }
}
```

There are differences between the results returned by medicalcoder and
comorbidity related to metastatic cancer and the ICD codes C7A.x.

```{r, echo = TRUE}
subset(charlson_delta, metacanc != mst)
subset(mdcr, patid %in% c(25628, 90045, 99058) & grepl("^C.[^\\d]", code))
```

Table 1 of @quan2005 defining the ICD codes mapping to metastatic solid tumor
as "C77.x--C80.x".   Under CMS codes there are C7A.x and C7B codes.  These codes
seem applicable to metastatic cancer.

```{r, echo = TRUE, results = "markup"}
subset(medicalcoder::get_icd_codes(with.descriptions = TRUE), grepl("^C(7[A-Z])", code))
```

The author of the comorbidity package is aware of this and has commented on
[GitHub](https://github.com/ellessenne/comorbidity/issues/16#issuecomment-513265712)

> I am a bit hesitant to add these codes to the scoring algorithm - after all, they did not exist at the time and they have not been validated in these settings (I think?). I think this is a much bigger "problem", as comorbidity codes evolve over time, but comorbidity scores barely do: how should researchers deal with this disconnection? ...I suppose there is no right or wrong answer to that! 😃

For medicalcoder, the choice was made to include these codes as C7A and C7B are,
in the hierachy, between C77 and C80.

This difference in approaches to mapping ICD codes to comorbidities accounts for
the differences in the metacanc/mst columns, and the score/cci colums.
```{r}
charlson_delta[["metacanc"]] <- NULL
charlson_delta[["mst"]] <- NULL
charlson_delta[["score"]] <- NULL
charlson_delta[["cci"]] <- NULL
```

The remaining columns in `charlson_delta` are:
```{r}
str(charlson_delta)
```

* `num_cmrb`: number of comorbidities flagged
* `cmrb_flag`: a 0/1 indicator for any comorbidity
* `age_score`: a score adjustment based on age _if_ age is provided.

```{r}
stopifnot(
  identical(
    names(charlson_delta),
    c("patid", "num_cmrb", "cmrb_flag", "age_score")
  )
)
```

An addition feature the medicalcoder package provides that the comorbidity
package does not, is a summary method for the results, which can be used to
build nice tables via
[kableExtra](https://CRAN.R-project.org/package=kableExtra).

```{r}
x <- summary(medicalcoder_charlson_results)[["conditions"]][, c("condition_description", "count", "percent")]

kbl(
  x = x,
  digits = 2,
  col.names = c("", "Count", "Percentage")
) |>
pack_rows("Comorbidity", start_row = 1, end_row = 17) |>
pack_rows("Number of Comorbidities", start_row = 18, end_row = nrow(x))
```

# Elixhauser Comorbidities

Applying the Elixhauser comorbidities to the `mdcr` data set via
`medicalcoder::comorbidities()` is done as follows.

```{r}
medicalcoder_elixhauser_results <-
  medicalcoder::comorbidities(
    data = mdcr,
    id.vars = "patid",
    icd.codes = "code",
    dx.var = "dx",
    icdv.var = "icdv",
    poa = 1L,       # assume all codes are present on admission
    primarydx = 0L, # assume all codes are secondary diagnoses
    method = "elixhauser_quan2005"
  )
```
Again, for the following comparisons we are only interested in the "conditions"
element of the return.
```{r}
str(medicalcoder_elixhauser_results, max.level = 1)
```

For `comorbidity::comorbidity()`, we will need to make two calls, one for the
ICD-9 codes and one for the ICD-10 codes.  We will then need to combine the
results and then apply a scoring function.

```{r}
comorbidity_elixhauser_icd9_results <-
  comorbidity::comorbidity(
    x = mdcr_icd9dx,
    id = "patid",
    code = "code",
    map = "elixhauser_icd9_quan",
    assign0 = TRUE # set less severe comorbidities flags to 0 when more severe comorbidities is also flagged
  )

comorbidity_elixhauser_icd10_results <-
  comorbidity::comorbidity(
    x = mdcr_icd10dx,
    id = "patid",
    code = "code",
    map = "elixhauser_icd10_quan",
    assign0 = TRUE
  )

# combine the ICD-9 and ICD-10 results into one set
comorbidity_elixhauser_results <-
  rbind(comorbidity_elixhauser_icd9_results, comorbidity_elixhauser_icd10_results)

comorbidity_elixhauser_results <-
  aggregate(. ~ patid, data = comorbidity_elixhauser_results, FUN = max)
```

We can now compare the comorbidity flags between the two packages.  First we
check that we have the same number of rows and all the patid are accounted for.

```{r}
nrow(medicalcoder_elixhauser_results[["conditions"]]) == nrow(comorbidity_elixhauser_results)

all(mdcr$patid %in% medicalcoder_elixhauser_results[["conditions"]][["patid"]])
all(medicalcoder_elixhauser_results[["conditions"]][["patid"]] %in% mdcr$patid)

all(mdcr$patid %in% comorbidity_elixhauser_results$patid)
all(comorbidity_elixhauser_results$patid %in% mdcr$patid)
```

There are differences in the results in terms of the names and numbers of
columns returned.

```{r}
dim(medicalcoder_elixhauser_results[["conditions"]])
names(medicalcoder_elixhauser_results[["conditions"]])

dim(comorbidity_elixhauser_results)
names(comorbidity_elixhauser_results)
```

We will compare results column by column.  First, we merge the data sets
together by patid.

```{r}
elixhauser_delta <-
  merge(
    x = comorbidity_elixhauser_results,
    y = medicalcoder_elixhauser_results[["conditions"]],
    all = TRUE,
    by = "patid"
  )
```

The relevant columns:

```{r, include = FALSE}
elixhauser_columns <- data.table::fread(text = "
Condition                                     | comorbidity | medicalcoder
AIDS/HIV                                      | aids        | AIDS
Alcohol abuse                                 | alcohol     | ALCOHOL
Anemias (Blood loss anaemia)                  | blane       | BLDLOSS
Anemias (Deficiency anaemia)                  | dane        | ANEMDEF
Cardiac Arrhythmias                           | carit       | CARDIAC_ARRHYTHMIAS
Cancer (Solid tummor without metastasis)      | solidtum    | TUMOR
Cancer (Metastatic)                           | metacanc    | METS
Chronic pulmonary disease                     | cpd         | CHRNLUNG
Coagulopathy                                  | coag        | COAG
Congestive Heart Failure                      | chf         | CHF
Depression                                    | depre       | DEPRESS
Diabetes (Complicated)                        | diabc       | DMCX
Diabetes (Uncomplicated)                      | diabunc     | DM
Drug abuse                                    | drug        | DRUG
Fluid and electrolyte disorders               | fed         | LYTES
Hypertension (Complicated)                    | hypc        | HTN_CX
Hypertension (Uncomplicated)                  | hypunc      | HTN_UNCX
Hypothyroidism                                | hypothy     | HYPOTHY
Liver disease                                 | ld          | LIVER
Lymphoma                                      | lymph       | LYMPH
Obesity                                       | obes        | OBESE
Other neurological disease                    | ond         | NEURO
Paralysis                                     | para        | PARA
Peptic ulcer disease excluding bleeding       | pud         | ULCER
Peripheral Vascular Disorders                 | pvd         | PERIVASC
Psychoses                                     | psycho      | PSYCH
Pulmonary Circulation Disorders               | pcd         | PULMCIRC
Renal Failure                                 | rf          | RENLFAIL
Rheumatoid artritis/collaged vascular disease | rheumd      | ARTH
Valvular disease                              | valv        | VALVE
Weight loss                                   | wloss       | WGHTLOSS
  ")
data.table::setDF(elixhauser_columns)
```

The following data.frame reports the condition and the corresponding column from
`comorbidity::comorbidity()` and from `medicalcoder::comorbidities()`.
```{r}
elixhauser_columns
```

For each row in `elixhauser_columns` we will verify that the results are
identical in `elixhauser_delta` and remove the columns.


```{r, include = FALSE}
# the HTN_C is only in the medicalcoder results
stopifnot(
  identical(
    medicalcoder_elixhauser_results[["conditions"]][["HTN_C"]],
    as.integer(
      medicalcoder_elixhauser_results[["conditions"]][["HTN_UNCX"]] |
        medicalcoder_elixhauser_results[["conditions"]][["HTN_CX"]]
    )
  )
)
```

```{r, echo = TRUE, results = "markup"}
for (i in seq_len(nrow(elixhauser_columns))) {
  x <- elixhauser_columns[["comorbidity"]][i]
  y <- elixhauser_columns[["medicalcoder"]][i]
  if (x == y) {
    x <- paste0(x, ".x")
    y <- paste0(y, ".y")
  }
  e <- base::substitute(identical(elixhauser_delta[[X]], elixhauser_delta[[Y]]), list(X = x, Y = y))
  print(e)
  r <- eval(e)
  print(r)
  stopifnot(r)
  elixhauser_delta[[x]] <- NULL
  elixhauser_delta[[y]] <- NULL
}
```

The remaining columns in the `elixhauser_delta` object are specific to
medicalcoder, save patid.

```{r}
str(elixhauser_delta)
```

```{r, include = FALSE}
stopifnot(identical(names(elixhauser_delta), c("patid", "HTN_C", "num_cmrb", "cmrb_flag", "mortality_index", "readmission_index")))
```

* `HTN_C`: _any_ hypertension - added to simplify the calculation of the
  mortality and readmission indices based on weights from AHRQ.

* `num_cmrb`:  Number of comorbidities flagged

* `cmrb_flag`: a 0/1 indicator for _any_ comorbidity being flagged

* `mortality_index` and `readmission_index`: index scores based on the AHRQ
  weights [@ahrq2017].

An addition feature of the medicalcoder package provides that the
comorbidity package does not, is a summary method for the results, which can
be used to build nice tables via
[kableExtra](https://CRAN.R-project.org/package=kableExtra).

```{r}
x <- summary(medicalcoder_elixhauser_results)[["conditions"]][, c("condition", "count", "percent")]

kbl(
  x = x,
  digits = 2,
  col.names = c("", "Count", "Percentage")
) |>
pack_rows("Comorbidity", start_row = 1, end_row = which(x$condition == ">= 1") - 1L) |>
pack_rows("Number of Comorbidities", start_row = which(x$condition == ">= 1"), end_row = nrow(x))
```

# All ICD Codes

Here we check for consistency in the mappings between _medicalcoder_ and
_comorbidity_ using all known ICD codes within _medicalcoder_.

```{r}
all_dx_codes <-
  subset(
    x      = medicalcoder::get_icd_codes(),
    subset = dx == 1 & !is.na(assignable_start),
    select = c("icdv", "full_code", "code")
    ) |>
  data.table::as.data.table() |>
  unique()
all_dx_codes[, code_id := paste0("ICD-", icdv, " ", full_code)]
```

## Charlson

```{r}
mdcr_results <-
  medicalcoder::comorbidities(
    data = all_dx_codes,
    icd.codes = "code",
    id.vars = "code_id",
    icdv.var = "icdv",
    method = "charlson_quan2005",
    poa = 1,
    primarydx = 0L)
mdcr_results <- mdcr_results[["conditions"]]

cmrb_results <-
  rbind(
    comorbidity::comorbidity(
      x = subset(all_dx_codes, icdv == 9),
      id = "code_id",
      code = "code",
      map = "charlson_icd9_quan",
      assign0 = TRUE
    ),
    comorbidity::comorbidity(
      x = subset(all_dx_codes, icdv == 10),
      id = "code_id",
      code = "code",
      map = "charlson_icd10_quan",
      assign0 = TRUE
    )
  )

charlson_delta <-
  merge(mdcr_results, cmrb_results, all = TRUE, by = "code_id")

for (i in seq_len(nrow(charlson_columns))) {
  x <- charlson_columns[["comorbidity"]][i]
  y <- charlson_columns[["medicalcoder"]][i]
  if (x == y) {
    x <- paste0(x, ".x")
    y <- paste0(y, ".y")
  }
  e <- base::substitute(identical(charlson_delta[[X]], charlson_delta[[Y]]), list(X = x, Y = y))
  print(e)
  r <- eval(e)
  print(r)
  if (r) {
    charlson_delta[[x]] <- NULL
    charlson_delta[[y]] <- NULL
  }
}

print(charlson_delta[mal != canc, .(code_id, mal, canc)], nrow = Inf)
```

```{r}
#| include: false
stopifnot(packageVersion("comorbidity") == "1.1.0")
stopifnot(charlson_delta[mal != canc, .N == 26L])
```

The ICD-10 codes C4A.x are not flagged as malignancies via
`comorbidity::comorbidity()`.  This is likely due to the mapping missing.  Here
we can see the mappings used between the two utilities.  The implementation used
by `comorbidity::comorbidity()` relies on regular expressions applied to the
data and the omitted C4A is why the code is not flagged.
`medicalcoder::comorbidities()` builds a table with valid ICD codes to join
against and has the C4A.x codes

```{r}
# comorbidity:::.maps$charlson_icd10_quan$canc
grep("^C4", comorbidity:::.maps$charlson_icd10_quan$canc, value = TRUE)
grep("^C4A", medicalcoder::get_charlson_codes()$code, value = TRUE)
grep("^C4A", medicalcoder::get_charlson_codes()$full_code, value = TRUE)
```

The implementation of `comorbidity::comorbidity()` can result in false positives
when invalid codes are submitted.

```{r}
df <- data.frame(id = "pat1", code = "C40-NOT-AN-ICD-CODE")
comorbidity::comorbidity(x = df, id = "id", code = "code", assign0 = TRUE, map = "charlson_icd10_quan")
medicalcoder::comorbidities(data = df, id.vars = "id", icd.codes = "code", method = "charlson_quan2005", poa = 1, primarydx = 0)[["conditions"]]
```

## Elixhauser
```{r}
mdcr_results <-
  medicalcoder::comorbidities(
    data = all_dx_codes,
    icd.codes = "code",
    id.vars = "code_id",
    icdv.var = "icdv",
    method = "elixhauser_quan2005",
    poa = 1,
    primarydx = 0L)
mdcr_results <- mdcr_results[["conditions"]]

cmrb_results <-
  rbind(
    comorbidity::comorbidity(
      x = subset(all_dx_codes, icdv == 9),
      id = "code_id",
      code = "code",
      map = "elixhauser_icd9_quan",
      assign0 = TRUE
    ),
    comorbidity::comorbidity(
      x = subset(all_dx_codes, icdv == 10),
      id = "code_id",
      code = "code",
      map = "elixhauser_icd10_quan",
      assign0 = TRUE
    )
  )

elixhauser_delta <-
  merge(
    x = cmrb_results,
    y = mdcr_results,
    all = TRUE,
    by = "code_id"
  )

data.table::setDT(elixhauser_delta)

for (i in seq_len(nrow(elixhauser_columns))) {
  x <- elixhauser_columns[["comorbidity"]][i]
  y <- elixhauser_columns[["medicalcoder"]][i]
  if (x == y) {
    x <- paste0(x, ".x")
    y <- paste0(y, ".y")
  }
  e <- base::substitute(identical(elixhauser_delta[[X]], elixhauser_delta[[Y]]), list(X = x, Y = y))
  print(e)
  r <- eval(e)
  print(r)
  if (r) {
    elixhauser_delta[[x]] <- NULL
    elixhauser_delta[[y]] <- NULL
  }
}
```

```{r, include = FALSE}
stopifnot(identical(names(elixhauser_delta), c("code_id", "HTN_C", "num_cmrb", "cmrb_flag", "mortality_index", "readmission_index")))
```

The only differences between the resulting data sets are columns unique to
medicalcoder.

```{r}
str(elixhauser_delta)
```

# Benchmarking

The medicalcoder package was built to use base R methods and has zero imports.
That said, if the input data set to `medicalcoder::comorbidities()` is a
[`data.table`](https://cran.r-project.org/package=data.table) and the
`data.table` namespace is available, the S3 methods for `data.table` will be used
and there will be a performance improvement. When given a
[`tibble`](https://tibble.tidyverse.org/) and the
[`dplyr`](https://dplyr.tidyverse.org/) namespace is available, the tibble-aware
path improves on base `data.frame` performance but remains slower than
`data.table`; see the project benchmarking results for details.

comorbidity imports several namespaces, including `data.table`, and uses
`data.table` for efficiency.

In the following benchmark we will look at the time required to apply the
Charlson comorbidity method to the `mdcr` data set via
`medicalcoder::comorbidities()` and `comorbidity::comorbidity()`. We will
consider the cases when inputting a data.frame and when inputting a data.table.
To support that work we make some copies of the input data and set the copies to
be `data.tables`.

<details>
  <summary>Code (click to toggle view/fold)</summary>

```{r}
mdcrDT <- data.table::copy(mdcr)
data.table::setDT(mdcrDT)
mdcr_icd9dxDT <- data.table::copy(mdcr_icd9dx)
mdcr_icd10dxDT <- data.table::copy(mdcr_icd10dx)
data.table::setDT(mdcr_icd9dxDT)
data.table::setDT(mdcr_icd10dxDT)
```
</details>

Next we define four functions to simplify calling the code multiple times.

<details>
  <summary>Code (click to toggle view/fold)</summary>

```{r}
medicalcoder_charlson_results <- function() {
  medicalcoder::comorbidities(
    data = mdcr,
    id.vars = "patid",
    icd.codes = "code",
    dx.var = "dx",
    icdv.var = "icdv",
    poa = 1L,
    primarydx = 0L,
    method = "charlson_quan2011"
  )
}

medicalcoder_charlson_results_DT <- function() {
  medicalcoder::comorbidities(
    data = mdcrDT,
    id.vars = "patid",
    icd.codes = "code",
    dx.var = "dx",
    icdv.var = "icdv",
    poa = 1L,
    primarydx = 0L,
    method = "charlson_quan2011"
  )
}

comorbidity_charlson_results <- function() {
  comorbidity_charlson_icd9_results <-
    comorbidity::comorbidity(
      x = mdcr_icd9dx,
      id = "patid",
      code = "code",
      map = "charlson_icd9_quan",
      assign0 = TRUE # set less severe comorbidities flags to 0 when more severe comorbidities is also flagged
    )

  comorbidity_charlson_icd10_results <-
    comorbidity::comorbidity(
      x = mdcr_icd10dx,
      id = "patid",
      code = "code",
      map = "charlson_icd10_quan",
      assign0 = TRUE
    )

# combine the ICD-9 and ICD-10 results into one set
  comorbidity_charlson_results <-
    rbind(comorbidity_charlson_icd9_results, comorbidity_charlson_icd10_results)

  comorbidity_charlson_results <-
    aggregate(. ~ patid, data = comorbidity_charlson_results, FUN = max)

# add the attributes to the combine set
  attributes(comorbidity_charlson_results)[c("class", "variable.labels", "map")] <-
    attributes(comorbidity_charlson_icd9_results)[c("class", "variable.labels", "map")]

  comorbidity_charlson_results[["score"]] <-
    comorbidity::score(
      x = comorbidity_charlson_results,
      weights = "quan",
      assign0 = TRUE
    )

# the score return is a numeric value, setting to integer to match the storage
# mode of medicalcoder
  comorbidity_charlson_results[["score"]] <-
    as.integer(comorbidity_charlson_results[["score"]])
}

comorbidity_charlson_results_DT <- function() {
  comorbidity_charlson_icd9_results <-
    comorbidity::comorbidity(
      x = mdcr_icd9dxDT,
      id = "patid",
      code = "code",
      map = "charlson_icd9_quan",
      assign0 = TRUE # set less severe comorbidities flags to 0 when more severe comorbidities is also flagged
    )

  comorbidity_charlson_icd10_results <-
    comorbidity::comorbidity(
      x = mdcr_icd10dxDT,
      id = "patid",
      code = "code",
      map = "charlson_icd10_quan",
      assign0 = TRUE
    )

# combine the ICD-9 and ICD-10 results into one set
  comorbidity_charlson_results <-
    rbind(comorbidity_charlson_icd9_results, comorbidity_charlson_icd10_results)

  comorbidity_charlson_results <-
    aggregate(. ~ patid, data = comorbidity_charlson_results, FUN = max)

# add the attributes to the combine set
  attributes(comorbidity_charlson_results)[c("class", "variable.labels", "map")] <-
    attributes(comorbidity_charlson_icd9_results)[c("class", "variable.labels", "map")]

  comorbidity_charlson_results[["score"]] <-
    comorbidity::score(
      x = comorbidity_charlson_results,
      weights = "quan",
      assign0 = TRUE
    )

# the score return is a numeric value, setting to integer to match the storage
# mode of medicalcoder
  comorbidity_charlson_results[["score"]] <-
    as.integer(comorbidity_charlson_results[["score"]])
}
```

</details>

Next we evaluate each function 20 times and record the total execution time, in
seconds, for each evaluation.

<details>
  <summary>Code (click to toggle view/fold)</summary>

```{r}
medicalcoder_times <- numeric(0)
medicalcoder_times_DT <- numeric(0)
comorbidity_times <- numeric(0)
comorbidity_times_DT <- numeric(0)

for (i in 1:20) {
  tic <- Sys.time()
  x <- medicalcoder_charlson_results()
  toc <- Sys.time()
  medicalcoder_times <- c(medicalcoder_times, as.numeric(difftime(toc, tic, units = "secs")))

  tic <- Sys.time()
  x <- medicalcoder_charlson_results_DT()
  toc <- Sys.time()
  medicalcoder_times_DT <- c(medicalcoder_times_DT, as.numeric(difftime(toc, tic, units = "secs")))

  tic <- Sys.time()
  x <- comorbidity_charlson_results()
  toc <- Sys.time()
  comorbidity_times <- c(comorbidity_times, as.numeric(difftime(toc, tic, units = "secs")))

  tic <- Sys.time()
  x <- comorbidity_charlson_results_DT()
  toc <- Sys.time()
  comorbidity_times_DT <- c(comorbidity_times_DT, as.numeric(difftime(toc, tic, units = "secs")))
}
```

</details>

The table below reports the summary of the time required to run each method.
`medicalcoder::comorbidities()` with a `data.table` input requires the least
time to compute. Due in part to the need to aggregate multiple calls for
`comorbidity::comorbidity()`, the call to `medicalcoder::comorbidities()` with a
`data.frame` is still much faster than `comorbidity::comorbidity()`.

```{r, echo = FALSE, results = "asis"}
cbind(medicalcoder_times_DT, medicalcoder_times, comorbidity_times_DT, comorbidity_times) |>
  apply(2, summary) |>
  t() |>
kbl(digits = 3) |>
kable_styling(bootstrap_options = "striped") |>
add_header_above(c("", "Time (seconds)" = 6))
```

# References

<!-- ----------------------------------------------------------------------- -->
<!--                              End of File                                -->
<!-- ----------------------------------------------------------------------- -->
