---
title: "medicalcoder vs icdcomorbid"
output:
 rmarkdown::html_vignette:
   toc: true
   number_sections: false
bibliography: ../references.bib
vignette: >
 %\VignetteIndexEntry{medicalcoder vs icdcomorbid}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r, label = "setup", include = FALSE}
# IMPORTANT SYNTAX NOTE:
# Vignettes cannot use the pipeOp `|>` so the package can depend on R >= 3.5.0
# The _articles_ (website only) can use the pipeOp since these are not part of the source package.

library(kableExtra)
options(qwraps2_markup = "markdown")
options(knitr.kable.NA = '')
knitr::opts_chunk$set(collapse = TRUE, fig.align = "center")
```

# Introduction

The purpose of this article is to compare the API and results between
medicalcoder and the R package
[icdcomorbid](https://cran.r-project.org/package=icdcomorbid)
[@r-icdcomorbid].

```{r}
library(medicalcoder)
library(icdcomorbid)
packageVersion("icdcomorbid")
packageDescription("icdcomorbid")$Title
cat(packageDescription("icdcomorbid")$Description)
```

icdcomorbid provides [@quan2005] Charlson and Elixhauser comorbidities.  A
nice feature of icdcomorbid is the ability to use custom mappings between
ICD codes and comorbidities.

# Prepare Data for icdcomorbid
The example data set `mdcr` within medicalcoder is in a format that is ideal
for `medicalcoder::comorbidities()`.  To prepare that data set for use in the
`icdcomorbid::icd9_to_comorbid()`  and `icdcomorbid::icd10_to_comorbid()`,
the data needs to be in a "wide" format instead of the provided long format.

We start by splitting the data into ICD-9 and ICD-10 sets and add rows for any
missing patids.

```{r}
mdcrDT <- data.table::as.data.table(medicalcoder::mdcr)
mdcrDT_icd9dx  <- subset(mdcrDT, icdv ==  9L & dx == 1L)
mdcrDT_icd10dx <- subset(mdcrDT, icdv == 10L & dx == 1L)
mdcrDT_icd9dx <-
  rbind(
    mdcrDT_icd9dx,
    data.table::data.table(
      patid = setdiff(mdcrDT$patid, mdcrDT_icd9dx$patid),
      icdv = 9L,
      code = "",
      dx = 1L
    )
  )
mdcrDT_icd10dx <-
  rbind(
    mdcrDT_icd10dx,
    data.table::data.table(
      patid = setdiff(mdcrDT$patid, mdcrDT_icd10dx$patid),
      icdv = 10L,
      code = "",
      dx = 1L
    )
  )
```

`icdcomorbid::long_to_wide()` is provided to transform data sets, but it is very
slow.  I waited about fifteen seconds for the following and then stopped the
process because there are more efficient ways.

```{r}
#  mdcrDT_icd9dx_wide <-
#    icdcomorbid::long_to_wide(
#      df = mdcrDT_icd9dx,
#      idx = "patid",
#      icd_cols = "code"
#    )
#
# killed process after 15 seconds
```

```{r}
tic <- Sys.time()

mdcrDT_icd9dx[, DX := paste0("DX", seq_along(code)), by = .(patid)]
mdcrDT_icd10dx[, DX := paste0("DX", seq_along(code)), by = .(patid)]

mdcrDT_icd9dx_wide <-
  data.table::dcast(mdcrDT_icd9dx, patid ~ DX, value.var = "code")

mdcrDT_icd10dx_wide <-
  data.table::dcast(mdcrDT_icd10dx, patid ~ DX, value.var = "code")

toc <- Sys.time()

difftime(toc, tic, units = "secs")
```

# Charlson Comorbidities

A detailed comparison between `medicalcoder::comorbidities()` and the methods in
the icdcomorbid package will not be presented here. Note the time required to
apply the Charlson comorbidities to the `mdcr` data set via
`medicalcoder::comorbidities()`.

```{r}
tic <- Sys.time()

medicalcoder_charlson_results <-
  medicalcoder::comorbidities(
    data      = mdcrDT,
    id.vars   = "patid",
    icd.codes = "code",
    dx.var    = "dx",
    icdv.var  = "icdv",
    poa       = 1L, # assume all codes are present on admission
    primarydx = 0L, # assume all codes are secondary diagnoses
    method    = "charlson_quan2011"
  )

toc <- Sys.time()

attr(medicalcoder_charlson_results, "tictoc") <-
  as.numeric(difftime(toc, tic, units = "secs"))
```

There are `r length(unique(mdcr$patid))` patids to assess.  To compare the
results between `medicalcoder::comorbidities()` and the results from the R
package icdcomorbid would require calling `icdcomorbid::icd9_to_comorbid()`,
`icd10_to_comorbid()`, aggregating the results, and scoring.

```{r, eval = FALSE}
tic <- Sys.time()

icdcomorbid_charlson_icd9_results <-
  icdcomorbid::icd9_to_comorbid(
    df = mdcrDT_icd9dx_wide,
    idx = "patid",
    icd_cols = grep("^DX", names(mdcrDT_icd9dx_wide), value = TRUE),
    mapping = "charlson9"
  )

icdcomorbid_charlson_icd10_results <-
  icdcomorbid::icd10_to_comorbid(
    df = mdcrDT_icd10dx_wide,
    idx = "patid",
    icd_cols = grep("^DX", names(mdcrDT_icd10dx_wide), value = TRUE),
    mapping = "charlson10"
  )

icdcomorbid_charlson_results <-
  rbind(
    data.table::setDT(icdcomorbid_charlson_icd9_results),
    data.table::setDT(icdcomorbid_charlson_icd10_results),
    use.names = TRUE,
    fill = TRUE
  )

# Slow aggregating
#icdcomorbid_charlson_results <-
#  aggregate(
#    . ~ patid,
#    data = icdcomorbid_charlson_results,
#    FUN = any,
#    na.rm = TRUE
#  )

# faster aggregating
icdcomorbid_charlson_results <-
  icdcomorbid_charlson_results[
    ,
    lapply(.SD, any),
    by = .(patid),
    .SDcols = -"patid"
  ]

toc <- Sys.time()

attr(icdcomorbid_charlson_results, "tictoc") <-
  as.numeric(difftime(toc, tic, units = "secs"))
```
```{r, include = FALSE, eval = FALSE}
saveRDS(
  object = icdcomorbid_charlson_results,
  file = "icdcomorbid_charlson_results.rds"
)
```
```{r, include = FALSE}
icdcomorbid_charlson_results <- 
  list(
    tryCatch(readRDS("vignettes/articles/icdcomorbid_charlson_results.rds"), warning = function(e) e),
    tryCatch(readRDS("articles/icdcomorbid_charlson_results.rds"), warning = function(e) e),
    tryCatch(readRDS("icdcomorbid_charlson_results.rds"), warning = function(e) e)
  )
icdcomorbid_charlson_results <- Filter(f = function(x) !(inherits(x, "warning")), icdcomorbid_charlson_results)
stopifnot(length(icdcomorbid_charlson_results) == 1L)
icdcomorbid_charlson_results <- icdcomorbid_charlson_results[[1]]
```

The amount of time required to process the data differs greatly between the
packages.
```{r}
attr(medicalcoder_charlson_results, "tictoc") # seconds
attr(icdcomorbid_charlson_results,  "tictoc") # seconds
```

The return from icdcomorbid are booleans. We will coerce to integers before
comparing to medicalcoder.

```{r}
data.table::setDT(icdcomorbid_charlson_results)
for (j in setdiff(names(icdcomorbid_charlson_results), "patid")) {
  data.table::set(
    icdcomorbid_charlson_results,
    j = j,
    value = as.integer(icdcomorbid_charlson_results[[j]])
  )
}
mdcr_v_icdcomorbid <-
  merge(
    x = medicalcoder_charlson_results,
    y = icdcomorbid_charlson_results,
    all = TRUE,
    by = c("patid")
  )
```

Let's compare the results starting the AIDS/HIV.
```{r}
identical(
  mdcr_v_icdcomorbid[["aidshiv"]],  # medicalcoder::comorbidities()
  mdcr_v_icdcomorbid[["aids_hiv"]]  # icdcomorbid
)

mdcr_v_icdcomorbid[aidshiv != aids_hiv, .(patid, aidshiv, aids_hiv)]

merge(
  x = data.table::as.data.table(mdcr)[patid %in% mdcr_v_icdcomorbid[aidshiv != aids_hiv, patid]],
  y = subset(
        x      = medicalcoder::get_charlson_codes(),
        subset = condition == "aidshiv",
        select = c("code", "icdv", "dx", "condition", "charlson_quan2005")
        ),
  by = c("code", "icdv", "dx")
)
```

There are differences in the flags for AIDS/HIV, all false-negatives from
icdcomorbid.  The issue, as shown below, is that ICD-9-CM code 042 is not
flagged as AIDS/HIV by icdcomorbid.  This appears to be due to the string
"042.x" using within the json files within icdcomorbid for mapping codes to
comorbidities.  Table 1 within Quan (2005) [@quan2005] reports "042.x-044.x"
which should match any ICD-9 code under the three digit code 042, 043, or 044.
(Note: 043 and 044 were removed from the ICD-9-CM standard starting October 1,
1994 (fiscal year 1995) which is before the earliest set of codes medicalcoder
has in its internal database.)

Within icdcomorbid, the references file for the Charlson ICD-9 codes appears to
have 042.x.  The false-negative suggests an issue with the way comorbidities are
flagged.

```{r}
scan(
  file = system.file(package = "icdcomorbid", "comorbidity_mappings", "charlson9.json"),
  what = "character",
  sep = "\n",
  quiet = TRUE
) |>
grep("aids_hiv", x = _, value = TRUE)
```

Next, congestive heart failure.
```{r}
identical(
  mdcr_v_icdcomorbid[["chf"]],  # medicalcoder::comorbidities()
  mdcr_v_icdcomorbid[["congestive_heart_failure"]]  # icdcomorbid
)
mdcr_v_icdcomorbid[
  chf != congestive_heart_failure,
  .(patid, chf, congestive_heart_failure)
]

mdcr_v_icdcomorbid[, .N, keyby = .(chf, congestive_heart_failure)]
```

Focusing on the differences:
```{r}
mdcr_v_icdcomorbid[
  !Vectorize(identical)(chf, congestive_heart_failure),
  .N,
  keyby = .(chf, congestive_heart_failure)
  ]
```

There are records which are missing form the results from icdcomorbid.  In these
cases, it appears that the missing rows are from records with no comorbidities
flagged. It should be noted that not all cases of `num_cmrb == 0` have missing
values for `congestive_heart_failure`.

```{r}
mdcr_v_icdcomorbid[num_cmrb == 0, .N]
mdcr_v_icdcomorbid[is.na(congestive_heart_failure), all(num_cmrb == 0)]
```

What about the cases where icdcomorbid flagged congestive_heart_failure and
medicalcoder did not
```{r}
#| label: icdcomorbid-chf-false-positives
# where does icdcomorbid flag that medicalcoder does not?
DT <-
  mdcrDT[patid %in% mdcr_v_icdcomorbid[congestive_heart_failure > chf, patid]]

all(DT$icdv == 10L)

DT <- DT[, unique(.SD), .SDcols = c("code")]
DT[, code_id := 1:.N]
DT[, DX := "DX"]
DT <- data.table::dcast(DT, code_id ~ DX, value.var = "code")

false_positives <-
  icdcomorbid::icd10_to_comorbid(
    df = DT,
    idx = "code_id",
    icd_cols = "DX", mapping = "charlson10"
  )
data.table::setDT(false_positives)

false_positives[(congestive_heart_failure), .N]

DT[
  false_positives[(congestive_heart_failure)], on = "code_id"
  ][
  , .(DX, congestive_heart_failure)
  ]
```

* P2930 (full code P29.30) should not be part of the Charlson comorbidities for
  CHF, I'm not sure why this is flagging at the moment, the JSON files within
  icdcomorbid report the code P29.0 only, which is correct and consistent with
  Table 1 of [@quan2005].

* I1310 (full code I13.10) should not flag based on what I see in the
  icdcomorbid JSON files.

```{r}
#| label: icdcomorbid-json-snippet
scan(
  file = system.file(package = "icdcomorbid", "comorbidity_mappings", "charlson10.json"),
  what = "character",
  sep = "\n",
  quiet = TRUE) |>
grep("congestive_heart_failure", x = _, value = TRUE)
```

And what about the cases where medicalcoder flags and icdcomorbid does not?
```{r}
#| label: icdcomorbid_false_negatives
icdcomorbid_false_negatives <-
  merge(
    x = mdcrDT[patid %in% mdcr_v_icdcomorbid[congestive_heart_failure < chf, patid]],
    y = subset(
          x      = medicalcoder::get_charlson_codes(),
          subset = condition == "chf",
          select = c("code", "icdv", "dx", "condition", "charlson_quan2005")
          ),
    by = c("code", "icdv", "dx")
  )
str(icdcomorbid_false_negatives)

unique(icdcomorbid_false_negatives[icdv == 9 & dx == 1, .(code)][["code"]])
unique(icdcomorbid_false_negatives[icdv == 10 & dx == 1, .(code)][["code"]])
```

These are all false negatives from icdcomorbid.

Given the false positive and false negative results, along with several orders
of magnitude longer computation time, we conclude here that medicalcoder is
preferable to icdcomorbid.

There are many more conditions to explore, but the above should be sufficient to
support using medicalcoder over icdcomorbid.



# References

<!-- ----------------------------------------------------------------------- -->
<!--                              End of File                                -->
<!-- ----------------------------------------------------------------------- -->
