---
title: "Transition From the _pccc_ Package to the _medicalcoder_ Package"
output:
 rmarkdown::html_vignette:
   toc: true
   number_sections: false
bibliography: ../references.bib
vignette: >
 %\VignetteIndexEntry{Transition From The pccc Package to the medicalcoder Package}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r, label = "setup", include = FALSE}
# IMPORTANT SYNTAX NOTE:
# Vignettes cannot use the pipeOp `|>` so the package can depend on R >= 3.5.0
# The _articles_ (website only) can use the pipeOp since these are not part of the source package.

library(kableExtra)
options(qwraps2_markup = "markdown")
options(knitr.kable.NA = '')
knitr::opts_chunk$set(collapse = TRUE, fig.align = "center")
```

# Introduction

The R package [_pccc_](https://cran.r-project.org/package=pccc)
[@feinstein2018pccc;@dewitt2026pccc] was published to support version 2 of
the Pediatric Complex Chronic Conditions (PCCC) [@feudtner2014pediatric].  This
document is provided to help users of _pccc_ to transition to _medicalcoder_.

Major differences between `pccc::ccc()` and `medicalcoder::comorbidities()`:

1. Input data format

   * `pccc::ccc()` expects a data.frame with each row representing one patient
     and/or encounter.  There is a column for each diagnostic and procedure code.
     For example, a dataset where the maximum number of diagnostic codes is six and
     the max number of procedure codes is five, an entry for patient XX could
     look like the following:

```
  patid    dx1    dx2    dx3    dx4 dx5 dx6     pr1     pr2     pr3     pr4 pr5
  patXX  T8619  E8809   E876  Z7982  NA  NA 02PAX3Z 5A1D70Z 04Q90ZZ 0TS60ZZ  NA
```


  * `medicalcoder::comorbidities()` expects the input data to be in a `data.frame`
    where each row is single ICD code.  For example, the same record for patient
    XX above would be three columns below, one row for each code, one column to
    identify the patient/encounter, and a column to denote if the code is a
    diagnostic (`dx = 1L`) or procedure (`dx = 0L`).

```
  patid    code  dx
  patXX   T8619   1
  patXX   E8809   1
  patXX    E876   1
  patXX   Z7982   1
  patXX 02PAX3Z   0
  patXX 5A1D70Z   0
  patXX 04Q90ZZ   0
  patXX 0TS60ZZ   0
```


2. ICD Version

  * `pccc::ccc()` considers ICD-9 and ICD-10 independently.  If the input
    data consists of both ICD-9 and ICD-10 data false negatives will be
    inevitable.  The version is set by the `icdv` argument to `pccc::ccc()`.

  * `medicalcoder::comorbidities()` considers both ICD-9 and ICD-10 at the same
    time.  A column added to the input data to identify the code version allows
    for a single patient/encounter record to consist of both version and to have
    PCCC flagged accordingly.  Users specify the ICD version via the arguments
    `icdv` and `icdv.var` to `medicalcoder::comorbidities()`.


3. PCCC Versions

  * `pccc::ccc()` only implements PCCC version 2 [@feudtner2014pediatric]

  * `medicalcoder::comorbidities()` implements:

    * `pccc_v2.0`: consistent results with `pccc::ccc()` for _pccc_ version 1.0.7.
    * `pccc_v2.1`: improved mappings of ICD codes to PCCC using the PCCC v2
      scoring algorithm.
    * `pccc_v3.0`: consistent with SAS code published with PCCC version 3
      [@feinstein2024pediatric].
    * `pccc_v3.1`: extended set of ICD code to condition mappings.

    * Note: _medicalcoder_ also provides several variants of the Charlson and
      Elixhauser comorbidities.

4. Subconditions

  * `pccc::ccc()` only returns flags for primary conditions

  * `medicalcoder::comorbidities()`: when the argument `subconditions = TRUE` is
    passed in, for PCCC the primary conditions and subconditions are flagged.
    Examples to follow.

5. Present-on-Admission and Longitudinal data

  * `pccc::ccc()` only considers singular encounters and considers all codes to
    be present-on-admission.

  * `medicalcoder::comorbidities()` can account for present-on-admission flags
    and longitudinal flagging of comorbidities within a patient over multiple
    encounters.

    For `flag.method = "cumulative"`, the final element of `id.vars` defines
    the within-patient encounter order and earlier ICD evidence can be carried
    forward.  This is especially relevant for PCCC v3 technology-dependent
    conditions, where a technology-dependent code may need to be evaluated
    together with non-technology-dependent evidence observed on another
    encounter.


# `pccc::ccc()` vs `medicalcoder::comorbidities()`

```{r, message=FALSE}
library(pccc)
packageVersion("pccc")
library(medicalcoder)
```

## Prepare Data

We'll use the `mdcr` dataset from the _medicalcoder_ package.

```{r}
head(mdcr)
```

We will split the dataset into two sets, one for ICD-9 and one for ICD-10.

Using the tidyverse packages, we can build the needed input datasets.

```{r}
mdcr_tbls <-
  mdcr |>
  dplyr::group_by(patid, icdv, dx) |>
  dplyr::mutate(n = seq_len(dplyr::n())) |>
  dplyr::ungroup() |>
  dplyr::mutate(dxv = dplyr::if_else(dx == 1, "dx", "pr")) |>
  dplyr::group_by(icdv) |>
  dplyr::group_split()
mdcr_tbls <-
  lapply(mdcr_tbls,
    tidyr::pivot_wider,
    id_cols = "patid",
    names_from = c("dxv", "n"),
    names_sep = "",
    values_from = "code"
  )
```

A `data.table` approach:

```{r}
mdcr_DTs <- data.table::as.data.table(data.table::copy(mdcr))
mdcr_DTs[
  ,
  dxv := paste0(data.table::fifelse(dx == 1, "dx", "pr"), seq_len(.N)),
  by = .(patid, icdv, dx)
]
mdcr_DTs <- split(mdcr_DTs, by = "icdv")
mdcr_DTs <-
  lapply(
    mdcr_DTs,
    data.table::dcast,
    formula = patid ~ dxv,
    value.var = "code",
    na.rm = FALSE
  )
```

## Applying `pccc::ccc()`

To flag PCCC via `pccc::ccc()` we need to call `pccc::ccc()` twice and to then
aggregate the results.

```{r}
tic <- Sys.time()

pccc_9_results_tbl <-
  pccc::ccc(
    data = mdcr_tbls[[1]],
    id = patid,
    dx_cols = grep("dx", names(mdcr_tbls[[1]]), value = TRUE),
    pc_cols = grep("pr", names(mdcr_tbls[[1]]), value = TRUE),
    icdv = 9L
  )

pccc_10_results_tbl <-
  pccc::ccc(
    data = mdcr_tbls[[2]],
    id = patid,
    dx_cols = grep("dx", names(mdcr_tbls[[2]]), value = TRUE),
    pc_cols = grep("pr", names(mdcr_tbls[[2]]), value = TRUE),
    icdv = 10L
  )

pccc_results_tbl <-
  dplyr::bind_rows(pccc_9_results_tbl, pccc_10_results_tbl) |>
  dplyr::group_by(patid) |>
  dplyr::summarize_all(max) |>
  dplyr::ungroup() |>
  dplyr::arrange(patid)

toc <- Sys.time()

pccc_ccc_tbl_time <- difftime(toc, tic, units = "secs")
```

```{r}
tic <- Sys.time()

pccc_9_results_DT <-
  pccc::ccc(
    data = mdcr_DTs[[1]],
    id = patid,
    dx_cols = grep("dx", names(mdcr_DTs[[1]]), value = TRUE),
    pc_cols = grep("pr", names(mdcr_DTs[[1]]), value = TRUE),
    icdv = 9L
  )

pccc_10_results_DT <-
  pccc::ccc(
    data = mdcr_DTs[[2]],
    id = patid,
    dx_cols = grep("dx", names(mdcr_DTs[[2]]), value = TRUE),
    pc_cols = grep("pr", names(mdcr_DTs[[2]]), value = TRUE),
    icdv = 10L
  )

pccc_results_DT <- data.table::rbindlist(list(pccc_9_results_DT, pccc_10_results_DT))

pccc_results_DT <-
  pccc_results_DT[, lapply(.SD, max), by = .(patid), .SDcols = -"patid"]
data.table::setkey(pccc_results_DT, patid)

toc <- Sys.time()

pccc_ccc_dt_time <- difftime(toc, tic)
```

A quick sanity check that we have the same results for both the tidyverse and
`data.table` input datasets.
```{r}
stopifnot(
  isTRUE(
    all.equal(pccc_results_DT, pccc_results_tbl, check.attributes = FALSE)
  )
)
```

## Calling `medicalcoder::comorbidities()`

`medicalcoder::comorbidities()` uses database-style joins of precomputed valid
ICD codes to conditions.  The `pccc::ccc()` implementation is based on partial
string matching.  The database-style joins used by
`medicalcoder::comorbidities()` requires less computational resources and time
to evaluate than `pccc::ccc()`.

```{r}
tic <- Sys.time()

medicalcoder_results <-
  medicalcoder::comorbidities(
    data      = mdcr,
    id.vars   = "patid",
    icd.codes = "code",
    icdv.var  = "icdv",
    dx.var    = "dx",
    method    = "pccc_v2.0",
    poa       = 1L # assuming all codes are POA
  )

toc <- Sys.time()
medicalcoder_df_time <- difftime(toc, tic)
```

## Differences in results?

```{r}
old_vs_new <-
  merge(
    x = pccc_results_DT,
    y = medicalcoder_results,
    all = TRUE,
    by = "patid",
    suffixes = c("_old", "_new")
  )
```

For the `mdcr` dataset, the any-condition flag (`ccc_flag` from `pccc::ccc()` and
`cmrb_flag` from `medicalcoder::comorbidities()`) are identical.

```{r}
stopifnot(isTRUE(with(old_vs_new, identical(ccc_flag, cmrb_flag))))
```

Second, the flags for all but the technology dependence and transplant flags are
identical.

```{r}
stopifnot(
  with(old_vs_new, identical(neuromusc_old,       neuromusc_new)),
  with(old_vs_new, identical(cvd_old,             cvd_new)),
  with(old_vs_new, identical(respiratory_old,     respiratory_new)),
  with(old_vs_new, identical(renal_old,           renal_new)),
  with(old_vs_new, identical(gi_old,              gi_new)),
  with(old_vs_new, identical(hemato_immu_old,     hemato_immu_new)),
  with(old_vs_new, identical(metabolic_old,       metabolic_new)),
  with(old_vs_new, identical(congeni_genetic_old, congeni_genetic_new)),
  with(old_vs_new, identical(malignancy_old,      malignancy_new)),
  with(old_vs_new, identical(neonatal_old,        neonatal_new))
)
```

Omitting the columns which are as expected from the `old_vs_new` data.table we
can focus in on the differences in the results.

```{r}
good <- c("neuromusc", "cvd", "respiratory", "renal", "gi", "hemato_immu",
  "metabolic", "congeni_genetic", "malignancy", "neonatal", "ccc_flag",
  "cmrb_flag")

for(g in good) {
  for (j in grep(g, names(old_vs_new), value = TRUE)) {
    data.table::set(old_vs_new, j = j, value = NULL)
  }
}
```
The remaining columns are:
```{r, include = FALSE}
# Sanity check
stopifnot(
  identical(
    names(old_vs_new),
    c("patid", "tech_dep", "transplant", "misc", "any_tech_dep", "any_transplant", "num_cmrb")
  )
)
```
```{r}
str(old_vs_new)
```

First, the `num_cmrb` column is a count of the number of conditions and is
reported by `medicalcoder::comorbidities()`. There is not similar flag from
`pccc::ccc()`.


```{r}
old_vs_new[, num_cmrb := NULL]
```

The `misc` column is the "miscellaneous" category reported by
`medicalcoder::comorbidities()` and is not reported by `pccc::ccc()`.  The
existence of the `misc` column and some differences in the returned results
between `pccc::ccc()` version 1.0.7, and `medicalcoder::comorbidities()` is due
to how _medicalcoder_ is implemented.

```{r}
old_vs_new
```

There are several ICD codes which need to be corrected in _pccc_

GitHub links:

* [ICD-9 349.1](https://github.com/CUD2V/pccc/issues/45)
* [ICD-9 V56](https://github.com/CUD2V/pccc/issues/46)
* [ICD-10 Z49](https://github.com/CUD2V/pccc/issues/47)
* [ICD-9 86.06](https://github.com/CUD2V/pccc/issues/48)
* [ICD-9 V45.85](https://github.com/CUD2V/pccc/issues/49)
* [ICD-9 V53.3](https://github.com/CUD2V/pccc/issues/50)
* [ICD-9 V53.91](https://github.com/CUD2V/pccc/issues/51)
* [ICD-9 V65.46](https://github.com/CUD2V/pccc/issues/52)
* [ICD-9 37.52](https://github.com/CUD2V/pccc/issues/53)
* [ICD-9 V42.0](https://github.com/CUD2V/pccc/issues/54)
* [ICD-10 Z94](https://github.com/CUD2V/pccc/issues/55)


# Additional Benefits of _medicalcoder_

## Computation Performance

_medicalcoder_ was built such that only base R is needed to install and use the
package. That said, there is specific support for tidyverse packages and
_data.table_. For example, the same calls as above but with either a `tibble` or
a `data.table` instead of a simple base R `data.frame` take less time to
compute. The differences here are small.  See
[benchmarking](https://github.com/dewittpe/medicalcoder/tree/main/benchmarking)
for more details.

```{r}
mdcr_tbl <- tibble::as_tibble(mdcr)
tic <- Sys.time()
medicalcoder_results <-
  medicalcoder::comorbidities(
    data      = mdcr_tbl,
    id.vars   = "patid",
    icd.codes = "code",
    icdv.var  = "icdv",
    dx.var    = "dx",
    method    = "pccc_v2.0",
    poa       = 1L
  )
toc <- Sys.time()
medicalcoder_tbl_time <- difftime(toc, tic, units = "secs")

mdcr_DT <- data.table::as.data.table(data.table::copy(mdcr))
tic <- Sys.time()
medicalcoder_results <-
  medicalcoder::comorbidities(
    data      = mdcr_DT,
    id.vars   = "patid",
    icd.codes = "code",
    icdv.var  = "icdv",
    dx.var    = "dx",
    method    = "pccc_v2.0",
    poa       = 1L
  )
toc <- Sys.time()
medicalcoder_dt_time <- difftime(toc, tic, units = "secs")
```

```{r}
pccc_ccc_tbl_time
pccc_ccc_dt_time
medicalcoder_df_time
medicalcoder_tbl_time
medicalcoder_dt_time
```

## Summary of results

A simple call to `summary()` will return a data.frame with counts and
percentages for the

```{r}
summary(medicalcoder_results)
```

## Subconditions

```{r, include = FALSE}
cvd_subconditions <- sort(unique(subset(get_pccc_codes(), condition == "cvd")$subcondition))
cvd_subconditions <- gsub("_", " ", cvd_subconditions)
cvd_subconditions <- paste0(seq_len(length(cvd_subconditions)), ". ", cvd_subconditions)
```

In the documentation for both PCCC v2 and v3 there are subconditions. For
example, there are `r length(cvd_subconditions)` subconditions under
cardiovascular disease:

```{r, echo = FALSE, results = 'asis'}
cat(cvd_subconditions, sep = "\n")
```

Calling `medicalcoder::comorbidities()` with `subconditions = TRUE` when working
with PCCC will flag these conditions as well as the primary conditions.

```{r}
with_subconditions <-
  medicalcoder::comorbidities(
    data          = mdcr,
    id.vars       = "patid",
    icd.codes     = "code",
    icdv.var      = "icdv",
    dx.var        = "dx",
    method        = "pccc_v2.0",
    poa           = 1L,
    subconditions = TRUE
  )

with_subconditions
```

The summary includes counts and percentages as before.  Additionally, for a
subcondition, the percentage is reported as percent of the cohort and as the
percent of those with the primary condition.

```{r}
str(summary(with_subconditions))
```

Using tools such as kableExtra, these summaries can be formatted into
publication ready tables.  For example, say we want to report on the
cardiovascular and metabolic conditions and subconditions.

```{r}
cvd_and_metabolic <- subset(summary(with_subconditions), condition %in% c("cvd", "metabolic"))
cvd_and_metabolic$subcondition[is.na(cvd_and_metabolic$subcondition)] <- "Any subcondition"

kableExtra::kbl(
  x = cvd_and_metabolic[, c("subcondition", "count", "percent_of_cohort", "percent_of_those_with_condition")],
  caption = "Patients with cardiovascular and/or metabolic conditions and the associated subconditions.",
  row.names = FALSE,
  digits = 2,
  col.names = c("Subcondition", "Patients", "% of cohort", "% of those with the primary condition")
) |>
kableExtra::kable_styling(bootstrap_options = "striped") |>
kableExtra::pack_rows(index = table(cvd_and_metabolic$condition))
```

# PCCC version 3
For more detail on the differences between PCCC v2 [@feudtner2014pediatric] and
PCCC v3 [@feinstein2024pediatric] see the
[PCCC article](https://www.peteredewitt.com/medicalcoder/articles/pccc.html#pccc-version-2-0-vs-pccc-version-3-0).

The manuscript uses PCCC v3.1 examples to show why encounter-level and
cumulative flagging can differ.  Under PCCC v3, technology-dependent ICD codes
do not always flag a condition by themselves; their interpretation can depend on
whether non-technology-dependent codes are present for the same encounter or
observed over the patient's record.  `medicalcoder::comorbidities()` exposes
that distinction with the same `flag.method` argument used for Charlson and
Elixhauser, while preserving a single long-format input structure.


# References

<!-- ----------------------------------------------------------------------- -->
<!--                              End of File                                -->
<!-- ----------------------------------------------------------------------- -->
