---
title: "Transition From The pccc Package to the medicalcoder Package"
output:
 rmarkdown::html_vignette:
   toc: true
   number_sections: false
bibliography: ../references.bib
vignette: >
 %\VignetteIndexEntry{Transition From The pccc Package to the medicalcoder Package}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r, label = "setup", include = FALSE}
# IMPORTANT SYNTAX NOTE:
# Vignettes cannot use the pipeOp `|>` so the package can depend on R >= 3.5.0
# The _articles_ (website only) can use the pipeOp since these are not part of the source package.

library(kableExtra)
options(qwraps2_markup = "markdown")
options(knitr.kable.NA = '')
knitr::opts_chunk$set(collapse = TRUE, fig.align = "center")
```

# Introduction

The R package [pccc](https://cran.r-project.org/package=pccc)
[@feinstein2018pccc;@dewitt2026pccc] was published to support version 2 of
the Pediatric Complex Chronic Conditions (PCCC) [@feudtner2014pediatric].  This
document is provided to help users of pccc to transition to medicalcoder.

Major differences between `pccc::ccc()` and `medicalcoder::comorbidities()`:

1. Input data format

   * `pccc::ccc()` expects a data.frame with each row representing one patient
     and/or encounter.  There is a column for each diagnostic and procedure code.
     For example, a data set were the max number of diagnostic codes is six and
     the max number of procedure codes is five, an entry for patient XX could
     look like the following:

```
  patid    dx1    dx2    dx3    dx4 dx5 dx6     pr1     pr2     pr3     pr4 pr5
  patXX  T8619  E8809   E876  Z7982  NA  NA 02PAX3Z 5A1D70Z 04Q90ZZ 0TS60ZZ  NA
```


  * `medicalcoder::comorbidities()` expects the input data to be in a data.frame
    where each row is single ICD code.  For example, the same record for patient
    XX above would be three columns below, one row for each code, one column to
    identify the patient/encounter, and a column to denote if the code is a
    diagnostic (dx = 1) or procedure (dx = 0).

```
  patid    code  dx
  patXX   T8619   1
  patXX   E8809   1
  patXX    E876   1
  patXX   Z7982   1
  patXX 02PAX3Z   0
  patXX 5A1D70Z   0
  patXX 04Q90ZZ   0
  patXX 0TS60ZZ   0
```


2. ICD Version

  * `pccc::ccc()` only considers ICD-9 and ICD-10 independently.  If the input
    data consists of both ICD-9 and ICD-10 data false negatives will be
    inevitable.  The version is set by the `icdv` argument to `pccc::ccc()`.

  * `medicalcoder::comorbidities()` considers both ICD-9 and ICD-10 at the same
    time.  A column added to the input data to identify the code version allows
    for a single patient/encounter record to consist of both version and to have
    PCCC flagged accordingly.  Users specify the ICD version via the arguments
    `icdv` and `icdv.var` to `medicalcoder::comorbidities()`.


3. PCCC Versions

  * `pccc::ccc()` only implements PCCC version 2 [@feudtner2014pediatric]

  * `medicalcoder::comorbidities()` implements:

    * `pccc_v2.0`: consistent results with `pccc::ccc()` for pccc version 1.0.7.
    * `pccc_v2.1`: improved mappings of ICD codes to PCCC using the PCCC v2
      scoring algorithm.
    * `pccc_v3.0`: consistent with SAS code published with PCCC version 3
      [@feinstein2024pediatric].
    * `pccc_v3.1`: extended set of ICD code to condition mappings.

    * Note: medicalcoder also provides several variants of the Charlson and
      Elixhauser comorbidities.

4. Subconditions

  * `pccc::ccc()` only returns flags for primary conditions

  * `medicalcoder::comorbidities()`: when the argument `subconditions = TRUE` is
    passed in, for PCCC the primary conditions and subconditions are flagged.
    Examples to follow.

5. Present-on-Admission and Longitudinal data

  * `pccc::ccc()` only considers singular encounters and considers all codes to
    be present-on-admission.

  * `medicalcoder::comorbidities()` can account for present-on-admission flags
    and longitudinal flagging of comorbidities within a patient over multiple
    encounters.


# `pccc::ccc()` vs `medicalcoder::comorbidities()`

```{r, message=FALSE}
library(pccc)
packageVersion("pccc")
library(medicalcoder)
```

## Prepare Data

We'll use the `mdcr` data set from the medicalcoder package.

```{r}
head(mdcr)
```

We will split the data set into two sets, one for ICD-9 and one for ICD-10.

Using the tidyverse we can build the needed input data sets

```{r}
mdcr_tbls <-
  mdcr |>
  dplyr::group_by(patid, icdv, dx) |>
  dplyr::mutate(n = seq_len(dplyr::n())) |>
  dplyr::ungroup() |>
  dplyr::mutate(dxv = dplyr::if_else(dx == 1, "dx", "pr")) |>
  dplyr::group_by(icdv) |>
  dplyr::group_split()
mdcr_tbls <-
  lapply(mdcr_tbls,
    tidyr::pivot_wider,
    id_cols = "patid",
    names_from = c("dxv", "n"),
    names_sep = "",
    values_from = "code"
  )
```

A data.table approach:

```{r}
mdcr_DTs <- data.table::as.data.table(data.table::copy(mdcr))
mdcr_DTs[
  ,
  dxv := paste0(data.table::fifelse(dx == 1, "dx", "pr"), seq_len(.N)),
  by = .(patid, icdv, dx)
]
mdcr_DTs <- split(mdcr_DTs, by = "icdv")
mdcr_DTs <-
  lapply(
    mdcr_DTs,
    data.table::dcast,
    formula = patid ~ dxv,
    value.var = "code",
    na.rm = FALSE
  )
```

## Applying `pccc::ccc()`

To flag PCCC via `pccc::ccc()` we need to call `pccc::ccc()` twice and to then
aggregate the results.

```{r}
tic <- Sys.time()

pccc_9_results_tbl <-
  pccc::ccc(
    data = mdcr_tbls[[1]],
    id = patid,
    dx_cols = grep("dx", names(mdcr_tbls[[1]]), value = TRUE),
    pc_cols = grep("pr", names(mdcr_tbls[[1]]), value = TRUE),
    icdv = 9
  )

pccc_10_results_tbl <-
  pccc::ccc(
    data = mdcr_tbls[[2]],
    id = patid,
    dx_cols = grep("dx", names(mdcr_tbls[[2]]), value = TRUE),
    pc_cols = grep("pr", names(mdcr_tbls[[2]]), value = TRUE),
    icdv = 10
  )

pccc_results_tbl <-
  dplyr::bind_rows(pccc_9_results_tbl, pccc_10_results_tbl) |>
  dplyr::group_by(patid) |>
  dplyr::summarize_all(max) |>
  dplyr::ungroup() |>
  dplyr::arrange(patid)

toc <- Sys.time()

pccc_ccc_tbl_time <- difftime(toc, tic, units = "secs")
```

```{r}
tic <- Sys.time()

pccc_9_results_DT <-
  pccc::ccc(
    data = mdcr_DTs[[1]],
    id = patid,
    dx_cols = grep("dx", names(mdcr_DTs[[1]]), value = TRUE),
    pc_cols = grep("pr", names(mdcr_DTs[[1]]), value = TRUE),
    icdv = 9
  )

pccc_10_results_DT <-
  pccc::ccc(
    data = mdcr_DTs[[2]],
    id = patid,
    dx_cols = grep("dx", names(mdcr_DTs[[2]]), value = TRUE),
    pc_cols = grep("pr", names(mdcr_DTs[[2]]), value = TRUE),
    icdv = 10
  )

pccc_results_DT <- data.table::rbindlist(list(pccc_9_results_DT, pccc_10_results_DT))

pccc_results_DT <-
  pccc_results_DT[, lapply(.SD, max), by = .(patid), .SDcols = -"patid"]
data.table::setkey(pccc_results_DT, patid)

toc <- Sys.time()

pccc_ccc_dt_time <- difftime(toc, tic)
```

A quick sanity check that we have the same results for both the tidyverse and
data.table input data sets.
```{r}
stopifnot(
  isTRUE(
    all.equal(pccc_results_DT, pccc_results_tbl, check.attributes = FALSE)
  )
)
```

## Calling `medicalcoder::comorbidities()`

```{r}
tic <- Sys.time()

medicalcoder_results <-
  medicalcoder::comorbidities(
    data = mdcr,
    id.vars = "patid",
    icd.codes = "code",
    icdv.var = "icdv",
    dx.var = "dx",
    method = "pccc_v2.0",
    poa = 1
  )

toc <- Sys.time()
medicalcoder_df_time <- difftime(toc, tic)
```

## Differences in results?

```{r}
old_vs_new <-
  merge(
    x = pccc_results_DT,
    y = medicalcoder_results[["conditions"]],
    all = TRUE,
    by = "patid",
    suffixes = c("_old", "_new")
  )
```

Most importantly, the condition flag (`ccc_flag` from `pccc::ccc()` and
`cmrb_flag` from `medicalcoder::comorbidities()`) are identical.

```{r}
stopifnot(
  isTRUE(
    with(old_vs_new, identical(ccc_flag, cmrb_flag))
  )
)
```

Second, the flags for all but the technology dependence and transplant flags are
identical.

```{r}
stopifnot(
  with(old_vs_new, identical(neuromusc_old,       neuromusc_new)),
  with(old_vs_new, identical(cvd_old,             cvd_new)),
  with(old_vs_new, identical(respiratory_old,     respiratory_new)),
  with(old_vs_new, identical(renal_old,           renal_new)),
  with(old_vs_new, identical(gi_old,              gi_new)),
  with(old_vs_new, identical(hemato_immu_old,     hemato_immu_new)),
  with(old_vs_new, identical(metabolic_old,       metabolic_new)),
  with(old_vs_new, identical(congeni_genetic_old, congeni_genetic_new)),
  with(old_vs_new, identical(malignancy_old,      malignancy_new)),
  with(old_vs_new, identical(neonatal_old,        neonatal_new))
)
```

Omitting the columns which are as expected from the `old_vs_new` data.table we
can focus in on the differences in the results.

```{r}
good <- c("neuromusc", "cvd", "respiratory", "renal", "gi", "hemato_immu",
  "metabolic", "congeni_genetic", "malignancy", "neonatal", "ccc_flag",
  "cmrb_flag")

for(g in good) {
  for (j in grep(g, names(old_vs_new), value = TRUE)) {
    data.table::set(old_vs_new, j = j, value = NULL)
  }
}

old_vs_new
```

First, the `num_cmrb` column is a count of the number of conditions and is
reported by `medicalcoder::comorbidities()`. There is not similar flag from
`pccc::ccc()`.


```{r}
old_vs_new[, num_cmrb := NULL]
```

The `misc` column is the "miscellaneous" category reported by
`medicalcoder::comorbidities()` and is not reported by `pccc::ccc()`.  The
existence of the `misc` column and some differences in the returned results
between `pccc::ccc()` version 1.0.7, and `medicalcoder::comorbidities()` is due
to how medicalcoder is implemented.

```{r}
old_vs_new
```

There are several ICD codes which need to be corrected in pccc

GitHub links:

* [ICD-9 349.1](https://github.com/CUD2V/pccc/issues/45)
* [ICD-9 V56](https://github.com/CUD2V/pccc/issues/46)
* [ICD-10 Z49](https://github.com/CUD2V/pccc/issues/47)
* [ICD-9 86.06](https://github.com/CUD2V/pccc/issues/48)
* [ICD-9 V45.85](https://github.com/CUD2V/pccc/issues/49)
* [ICD-9 V53.3](https://github.com/CUD2V/pccc/issues/50)
* [ICD-9 V53.91](https://github.com/CUD2V/pccc/issues/51)
* [ICD-9 V65.46](https://github.com/CUD2V/pccc/issues/52)
* [ICD-9 37.52](https://github.com/CUD2V/pccc/issues/53)
* [ICD-9 V42.0](https://github.com/CUD2V/pccc/issues/54)
* [ICD-10 Z94](https://github.com/CUD2V/pccc/issues/55)


# Additional Benefits of medicalcoder

## Computation Performance

medicalcoder was built such that only base R is needed to install and use the
package.  That said, there is specific support for the tidyverse and data.table.
For example, the same calls as above but with either a tibble or a data.table
instead of a simple base R data.table take less time to compute.  The
differences here are small.  See
[benchmarking](https://github.com/dewittpe/medicalcoder/tree/main/benchmarking)
for more details.

```{r}
mdcr_tbl <- tibble::as_tibble(mdcr)
tic <- Sys.time()
medicalcoder_results <-
  medicalcoder::comorbidities(
    data = mdcr_tbl,
    id.vars = "patid",
    icd.codes = "code",
    icdv.var = "icdv",
    dx.var = "dx",
    method = "pccc_v2.0",
    poa = 1
  )
toc <- Sys.time()
medicalcoder_tbl_time <- difftime(toc, tic, units = "secs")

mdcr_DT <- data.table::as.data.table(data.table::copy(mdcr))
tic <- Sys.time()
medicalcoder_results <-
  medicalcoder::comorbidities(
    data = mdcr_DT,
    id.vars = "patid",
    icd.codes = "code",
    icdv.var = "icdv",
    dx.var = "dx",
    method = "pccc_v2.0",
    poa = 1
  )
toc <- Sys.time()
medicalcoder_dt_time <- difftime(toc, tic, units = "secs")
```

```{r}
pccc_ccc_tbl_time
pccc_ccc_dt_time
medicalcoder_df_time
medicalcoder_tbl_time
medicalcoder_dt_time
```

## Summary of results

A simple call to `summary()` returns a data frame with counts and percentages
for the PCCC conditions.

```{r}
summary(medicalcoder_results)
```

## Subconditions

```{r, include = FALSE}
cvd_subconditions <- sort(unique(subset(get_pccc_codes(), condition == "cvd")$subcondition))
cvd_subconditions <- gsub("_", " ", cvd_subconditions)
cvd_subconditions <- paste0(seq_len(length(cvd_subconditions)), ". ", cvd_subconditions)
```

In the documentation for both PCCC v2 and v3 there are subconditions. For
example, there are `r length(cvd_subconditions)` subconditions under
cardiovascular disease:

```{r, echo = FALSE, results = 'asis'}
cat(cvd_subconditions, sep = "\n")
```

Calling `medicalcoder::comorbidities()` with `subconditions = TRUE` when working
with PCCC will flag these conditions as well as the primary conditions.

```{r}
with_subconditions <-
  medicalcoder::comorbidities(
    data = mdcr,
    id.vars = "patid",
    icd.codes = "code",
    icdv.var = "icdv",
    dx.var = "dx",
    method = "pccc_v2.0",
    poa = 1,
    subconditions = TRUE
  )

str(with_subconditions, max.level = 1)
```

The summary includes counts and percentages as before.  Additionally, for a
subconditon, the percentage is reported as percent of the cohort and as the
percent of those with the primary condition.

```{r}
str(summary(with_subconditions))
```

Using tools such as kableExtra, these summaries can be formatted into
publication ready tables.  For example, say we want to report on the
cardiovascular and metabolic conditions and subconditions.

```{r}
cvd_and_metabolic <- subset(summary(with_subconditions), condition %in% c("cvd", "metabolic"))
cvd_and_metabolic$subcondition[is.na(cvd_and_metabolic$subcondition)] <- "Any subcondition"

kableExtra::kbl(
  x = cvd_and_metabolic[, c("subcondition", "count", "percent_of_cohort", "percent_of_those_with_condition")],
  caption = "Patients with cardiovascular and/or metabolic conditions and the associated with_subconditions.",
  row.names = FALSE,
  digits = 2,
  col.names = c("Subcondition", "Patients", "% of chort", "% of those with the primary condition")
) |>
kableExtra::kable_styling(bootstrap_options = "striped") |>
kableExtra::pack_rows(index = table(cvd_and_metabolic$condition))
```

# PCCC version 3
For more detail on the differences between PCCC v2 [@feudtner2014pediatric] and
PCCC v3 [@feinstein2024pediatric] see the
[PCCC article](https://www.peteredewitt.com/medicalcoder/articles/pccc.html#pccc-version-2-0-vs-pccc-version-3-0).


# References

<!-- ----------------------------------------------------------------------- -->
<!--                              End of File                                -->
<!-- ----------------------------------------------------------------------- -->
